### 2024-11-29 paper/main.tex
Paper: WIP

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[thinc]{esdiff}

\begin{document}

\title{Parallel-in-time methods for Particle Simulations}

\author{\IEEEauthorblockN{Lukáš Veškrna}
\IEEEauthorblockA{\textit{Technical University Munich} \\
Munich, Germany \\
lukas.veskrna@gmail.com}
}

\maketitle

\begin{abstract}
TBD
\end{abstract}

\begin{IEEEkeywords}
parareal, particle-simulations, parallel-in-time.
\end{IEEEkeywords}

\section{Introduction}

Particle simulations are an important tool for understanding many different kinds of particle-based systems, such as molecular, fluid, plasma, solar system and stellar dynamics.

Simulating those systems usually requires huge computations, which can take a very long time. Some of these computations can be run in parallel in order to reduce the computation time. There has been more focus in parallelizing these simulations through the spatial domain, which allows us to run bigger simulations in the same time.

There do, however, exist algorithms that allow us to parallelize simulations in the time domain. This could be very useful for running smaller simulations through a very long timespan. This is of interest for example in solar system dynamics. If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time. Thus, parallel-in-time algorithms can be employed.\cite{parallelsolar}

In particle simulations, we are often numerically solving an ODE in the form:
$$
\diff{u(t)}{t} = f(t, u(t))
$$
where $f$ is the function describing our problem, $u(t)$ is the state of the system at time $t$ and we are given an initial condition $u(t_0) = u_0$.

The integration part is the part which usually is dependent on some previous time step and thus can't be easily parallelized. In order to solve this problem, multiple techniques were devised with the most widely known one being the parareal algorithm.

\section{The parareal algorithm}

A description of the parareal algorithm is given in \cite{parareal}. It achieves parallel-in-time integration using two types of integrators. A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to fill in the gaps in this estimated trajectory.

Supposing we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta T = \frac{T-t_0}{N}$. We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta T$ (so $t_0 = t_0$ and $t_N = T$).

The algorithm works in iterations. The 0th iteration is used to initialize the algorithm, while the subsequent iterations improve the result of the previous iteration.

In the 0th iteration, we use the coarse solver to compute $u_1\dots u_n$ sequentially. For every $0 \leq n < N$ we do
$$
u^0_{n+1} = \mathcal{C}(t_n, t_n+\Delta T, u^0_n)
$$
with $u^0_0 = u_0$, which is the initial state of the system.

In every subsequent iteration $i+1$, we first calculate the results of the fine but expensive solver, $\mathcal{F}(t_n, t_n+\Delta T, u^{i}_n)$, for every $0 \leq n < N$ in parallel.

Then, we adjust the estimated states from the previous iteration with the following correction:
\begin{multline}
u^{i+1}_{n+1} = \mathcal{C}(t_n, t_n+\Delta T, u^{i+1}_n)\\ + \mathcal{F}(t_n, t_n+\Delta T, u^i_n)\\ - \mathcal{C}(t_n, t_n+\Delta T, u^{i}_n)
\end{multline}

This is repeated until satisfactory convergence. It has been proven for example by \cite{parareal} (See Proposition 1) that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.

This method has been formulated using different generalized frameworks in \cite{parareal2}. They also extend the time interval paritioning to arbitrary partions, not only equally sized ones as in \cite{parareal}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_1.eps}}
\caption{Parareal convergence for the example of a harmonic oscillator. The coarse solver is Euler and the fine solver is RK4}
\label{fig}
\end{figure}

\section{Application to particle simulations}
The purpose of this paper is to examine the potential in the use of the parareal method for long running simulations such as simulations of the solar system.

\subsection{The particle simulation}
We use a simple $n$-body particle simulation of the solar system with time complexity $\mathcal{O}(n^2)$, which we then integrate using the parareal method. Thus, we are integrating the following function:
$$
f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
$$
where 
$$
a_i(t) = -G \sum_j m_j\frac{x_i(t)-x_j(t)}{\|x_i(t)-x_j(t)\|_2^3}
$$
where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from the Newton's law of universal gravitation.

\subsection{Choice of the coarse and fine integrators}
\begin{itemize}
    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its better precision. 
    \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic. A modified parareal algorithm that preserves the symplectic property, shown in \cite{symplecticparareal}, also exists, however it is beyond the scope of this paper.
    \item An adaptive step size fine integrator with error control would be an option, such as the Dormand-Prince RK45 method. However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
\end{itemize}

\section{Implementation}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_2.eps}}
\caption{Path of Mercury over the time of one Earth year in the simulation converging over multiple iterations of the parareal algorithm. Velocity verlet and RK4 are used with the same step size of $0.5\text{day}$.}
\label{fig}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
\caption{\color{red} TODO}
\label{fig}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{\color{red} TODO}
\label{fig}
\end{figure}

\begin{thebibliography}{00}
\bibitem{parareal}Bal, G. \& Maday, Y. A “Parareal” Time Discretization for Non-Linear PDE's with Application to the Pricing of an American Put. {\em Recent Developments In Domain Decomposition Methods}. pp. 189-202 (2002)
6-578 (2007), [Online]. Available: https://doi.org/10.1007/978-3-642-56118-4\_12
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), http://dx.doi.org/10.1086/118485
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: https://doi.org/10.1137/05064607X
\end{thebibliography}

\end{document}

### 2024-11-29 paper/fig_1.py

@@ -1,0 +1,17 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+
+import sys
+sys.path.append("..")
+
+from ref.parareal import test_problem
+
+
+def fig_1():
+	test_problem(show=False)
+	plt.tight_layout()
+	plt.savefig("fig_1.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_1()

### 2024-11-29 paper/fig_2.py

@@ -1,0 +1,44 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+import numpy as np
+
+import sys
+sys.path.append("..")
+
+from ref.parareal import parareal, verlet, rk4
+from ref.simulation import simulation
+import ref.planets as planets
+
+
+def fig_2():
+	fig = plt.figure(figsize=plt.figaspect(0.5))
+
+	it = 0
+
+	def plot_one(states):
+		nonlocal it
+		ax = fig.add_subplot(2, 3, it+1, projection='3d')
+		ax.get_xaxis().set_ticks([])
+		ax.get_yaxis().set_ticks([])
+		ax.get_zaxis().set_ticks([])
+		ax.title.set_text(f"iter {it}")
+
+		ax.plot3D(states[:, 0, 1, 0], states[:, 0, 1, 1], states[:, 0, 1, 2], label="mercury", color="orange")
+		it += 1
+
+	initial = np.array([planets.positions, planets.velocities])
+	ts, states = parareal(
+		simulation, 0, 365, 
+		initial, 
+		num_segments=365*2, 
+		max_iters=6, 
+		eps=1e-100, 
+		coarse_solver=verlet, fine_solver=rk4,
+		observer=plot_one)
+	
+	plt.tight_layout()
+	plt.savefig("fig_2.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_2()

### 2024-11-29 paper/fig_3.py

@@ -1,0 +1,34 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+import numpy as np
+
+import sys
+sys.path.append("..")
+import ref.planets as planets
+
+
+def fig_3():
+	data = np.loadtxt("parareal_test_out.txt.7069723")
+	data, energy = data[:, :-2], data[:, -2:]
+
+	data = data.reshape((data.shape[0], data.shape[1]//3, 3))
+
+	lim = 1.2*max(data[0, :, :].min(), data[0, :, :].max())
+
+	plt.rcParams["grid.color"] = "#DDD"
+	ax = plt.axes(projection="3d")
+	ax.set_xlim3d(-lim, lim)
+	ax.set_ylim3d(-lim, lim)
+	ax.set_zlim3d(-lim, lim)
+	plt.xlabel("x [AU]")
+	plt.ylabel("y [AU]")
+
+	for i, name in enumerate(planets.names):
+		ax.plot3D(data[:, i, 0], data[:, i, 1], data[:, i, 2], label=name.capitalize())
+	ax.legend(bbox_to_anchor=(0.1, 1), loc="upper left", bbox_transform=plt.gcf().transFigure)
+
+	plt.savefig("fig_3.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_3()

### 2024-11-29 paper/fig_4.py

@@ -1,0 +1,30 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+plt.rcParams.update({
+	"text.usetex": True,
+})
+import numpy as np
+
+import sys
+sys.path.append("..")
+import ref.planets as planets
+
+
+def fig_4():
+	data = np.loadtxt("parareal_test_out.txt.7069723")
+	data, energy = data[:, :-2], data[:, -2:]
+
+	ax = plt.axes()
+	ax.plot(np.arange(len(energy))*100, energy[:, 0], label="kinetic energy")
+	ax.plot(np.arange(len(energy))*100, energy[:, 1], label="potential energy")
+	ax.plot(np.arange(len(energy))*100, energy.sum(axis=1), label="total energy")
+	ax.legend()
+	plt.xlabel("Time [day]")
+	plt.ylabel(r"Energy [$\textrm{M}_{\textrm{Earth}}\frac{\textrm{AU}^2}{\textrm{day}^2}$]")
+
+	plt.tight_layout()
+	plt.savefig("fig_4.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_4()

### 2024-11-30 paper/main.tex
Job edits

@@
 \usepackage{graphicx}
+\usepackage{booktabs}
 \usepackage{textcomp}
@@
 \caption{Parareal convergence for the example of a harmonic oscillator. The coarse solver is Euler and the fine solver is RK4}
-\label{fig}
+\label{oscillator}
 \end{figure}
@@
 Velocity verlet and RK4 are used with the same step size of $0.5\text{day}$.}
-\label{fig}
+\label{mercury}
 \end{figure}
@@
 \caption{\color{red} TODO}
-\label{fig}
+\label{longrunning}
 \end{figure}
@@
 \caption{\color{red} TODO}
-\label{fig}
+\label{energy}
 \end{figure}
+
+\begin{table*}[htbp]
+\caption{Comparison of running times taken by different configurations of the simulator}
+\begin{center}
+\begin{tabular}{ccrrrrrrcc}
+\toprule
+\textbf{Program} & \textbf{Environment} & \textbf{n CPUs} & \textbf{Segments} & \textbf{Fine steps} & \textbf{Coarse steps} & \textbf{Final steps} & \textbf{CPU time} & \textbf{Runtime} \\
+\midrule
+parareal & MetaCentrum & 30 & 30 & $10^5$ & 100 & $30\cdot 10^5$ & 02:01:57 & 00:04:30 \\
+serial   & MetaCentrum &  1 &  1 & N/A    & N/A & $30\cdot 10^5$ & 00:06:27 & 00:06:27 \\
+parareal & Laptop      &  4 &  4 & $10^5$ & 100 & $4\cdot 10^5$ & 00:03:26 & 00:00:51 \\
+serial   & Laptop      &  1 &  1 & N/A    & N/A & $4\cdot 10^5$ & 00:00:30 & 00:00:30 \\
+\bottomrule
+\end{tabular}
+\label{runtimes}
+\end{center}
+\end{table*}
 

### 2024-11-30 parareal/dist.py

@@ -1,0 +1,20 @@
+import numpy as np
+import sys
+
+
+if len(sys.argv) <= 2:
+	print("not enough arguments", file=sys.stderr)
+	sys.exit(1)
+
+data_1 = np.loadtxt(sys.argv[1])
+data_1, energy_1 = data_1[:, :-2], data_1[:, -2:]
+data_1 = data_1.reshape((data_1.shape[0], data_1.shape[1]//3, 3))
+
+data_2 = np.loadtxt(sys.argv[2])
+data_2, energy_2 = data_2[:, :-2], data_2[:, -2:]
+data_2 = data_2.reshape((data_2.shape[0], data_2.shape[1]//3, 3))
+
+data_1 = data_1[:min(len(data_1), len(data_2))]
+data_2 = data_2[:min(len(data_1), len(data_2))]
+
+print(np.abs(data_1-data_2).max())

### 2024-11-30 parareal/job.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
 #PBS -N parareal_test
-#PBS -l select=1:ncpus=30:ompthreads=30:mem=1gb:scratch_local=1gb
-#PBS -l walltime=00:05:00
+#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb
+#PBS -l walltime=00:10:00
 
 export OMP_NUM_THREADS=$PBS_NUM_PPN
 
@@ -10,25 +10,24 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # append a line to a file "jobs_info.txt" containing the ID of the job, the hostname of the node it is run on, and the path to a scratch directory
 # this information helps to find a scratch directory in case the job fails, and you need to remove the scratch directory manually 
-echo "$PBS_JOBID is running on node `hostname -f` in a scratch directory $SCRATCHDIR" >> $DATADIR/jobs_info.txt
-
-# (old) load openmpi modules
-# module add openmpi
-
-# test if the scratch directory is set
-# if scratch directory is not set, issue error message and exit
-test -n "$SCRATCHDIR" || { echo >&2 "Variable SCRATCHDIR is not set!"; exit 1; }
-
-# copy program
-cp $DATADIR/build/parareal $SCRATCHDIR || { echo >&2 "Error while copying input file(s)!"; exit 2; }
-cd $SCRATCHDIR
-chmod 750 parareal
+echo "$PBS_JOBID is running on node `hostname -f`" >> $DATADIR/jobs_info.txt
 
 # run program
-./parareal -i $DATADIR/planets.cfg -o out.txt --nseg $OMP_NUM_THREADS --coarse-steps 100 --fine-steps 100000 --time 365000 --log-period 100 || { echo >&2 "Calculation ended up erroneously (with a code $?) !!"; exit 3; }
+NSEG=32
+COARSE_STEPS=5000
+FINE_STEPS=50000
+TIME=10000
+EPS=1e-3
+MAX_ITERS=16
+LOG_PERIOD=10
 
-# move the output to user's DATADIR or exit in case of failure
-cp out.txt $DATADIR/ || { echo >&2 "Result file(s) copying failed (with a code $?) !!"; exit 4; }
+echo "=== parareal ==="
[…]

### 2024-11-30 parareal/serial.sh

@@ -1,34 +1,0 @@
-#!/bin/bash
-#PBS -N parareal_test
-#PBS -l select=1:ncpus=1:ompthreads=1:mem=500mb:scratch_local=500mb
-#PBS -l walltime=00:15:00
-
-export OMP_NUM_THREADS=$PBS_NUM_PPN
-
-# define a DATADIR variable: directory where the input files are taken from and where the output will be copied to
-DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
-
-# append a line to a file "jobs_info.txt" containing the ID of the job, the hostname of the node it is run on, and the path to a scratch directory
-# this information helps to find a scratch directory in case the job fails, and you need to remove the scratch directory manually 
-echo "$PBS_JOBID is running on node `hostname -f` in a scratch directory $SCRATCHDIR" >> $DATADIR/jobs_info.txt
-
-# (old) load openmpi modules
-# module add openmpi
-
-# test if the scratch directory is set
-# if scratch directory is not set, issue error message and exit
-test -n "$SCRATCHDIR" || { echo >&2 "Variable SCRATCHDIR is not set!"; exit 1; }
-
-# copy program
-cp $DATADIR/build/parareal $SCRATCHDIR || { echo >&2 "Error while copying input file(s)!"; exit 2; }
-cd $SCRATCHDIR
-chmod 750 parareal
-
-# run program
-./parareal -i $DATADIR/planets.cfg -o out.txt --nseg 30 --coarse-steps 100 --fine-steps 100000 --time 365000 --log-period 100 --serial || { echo >&2 "Calculation ended up erroneously (with a code $?) !!"; exit 3; }
-
-# move the output to user's DATADIR or exit in case of failure
-cp out.txt $DATADIR/ || { echo >&2 "Result file(s) copying failed (with a code $?) !!"; exit 4; }
-
-# clean the SCRATCH directory
-clean_scratch

### 2024-11-30 parareal/src/main.cpp

@@ -23,7 +23,7 @@ int main(int argc, char** argv) {
 	        ("coarse-steps", "Number of steps of the coarse integrator", cxxopts::value<std::size_t>())
 	        ("fine-steps", "Number of steps of the coarse integrator", cxxopts::value<std::size_t>())
 	        ("max-iters", "Maximum number of parareal iterations", cxxopts::value<std::size_t>()->default_value("1000"))
-	        ("eps", "Epsilon for parareal convergence", cxxopts::value<double>()->default_value("1e-12"))
+	        ("eps", "Epsilon for parareal convergence", cxxopts::value<double>()->default_value("1e-5"))
 	        ("vis", "Boolean", cxxopts::value<bool>()->default_value("false"))
 	        ("serial", "Switch to serial algorithm (for comparison)", cxxopts::value<bool>()->default_value("false"))
 	        ("log-period", "Log every `log_period` timesteps", cxxopts::value<std::size_t>()->default_value("1"))

### 2024-11-30 parareal/test.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+cd build
+cmake --build .
+
+NSEG=32
+COARSE_STEPS=5000
+FINE_STEPS=50000
+TIME=10000
+EPS=1e-3
+MAX_ITERS=8
+LOG_PERIOD=10
+
+echo "=== parareal ==="
+time ./parareal -i ../planets.cfg -o out_p.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
+echo "===  serial  ==="
+time ./parareal --serial -i ../planets.cfg -o out.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
+
+echo "=== distance ==="
+python3 ../dist.py out_p.txt out.txt
+
+#python3 vis.py out_p.txt

### 2024-11-30 parareal/dist.py
Better error measuring

@@ -1,9 +1,11 @@
+import matplotlib.pyplot as plt
 import numpy as np
 import sys
 
 
 if len(sys.argv) <= 2:
 	print("not enough arguments", file=sys.stderr)
+	print("python3 dist.py <parareal out> <serial out>")
 	sys.exit(1)
 
 data_1 = np.loadtxt(sys.argv[1])
@@ -17,4 +19,19 @@ data_2 = data_2.reshape((data_2.shape[0], data_2.shape[1]//3, 3))
 data_1 = data_1[:min(len(data_1), len(data_2))]
 data_2 = data_2[:min(len(data_1), len(data_2))]
 
-print(np.abs(data_1-data_2).max())
+abs_err = np.linalg.norm(data_1-data_2, axis=-1)
+
+print("max abs err: ", abs_err.max())
+print("mean abs err: ", abs_err.mean())
+
+tolerance = 0.1
+rel_errs = abs_err/np.linalg.norm(data_2, axis=-1)
+print("max rel err: ", rel_errs.max())
+print("mean rel err: ", rel_errs.mean())
+
+print("no. of outliers (rel err > 0.1): ", (rel_errs > tolerance).sum() / len(data_1))
+
+plt.hist(rel_errs[:, 1])
+
+#plt.plot(np.arange(len(data_1)), rel_errs[:, 1])
+plt.show()

### 2024-11-30 paper/main.tex
Correct results in table

--- 2024-11-29 paper/main.tex
@@
 \end{IEEEkeywords}
+
+\section*{Acknowledgment}
+\addcontentsline{toc}{section}{Acknowledgment}
+
+I am greatful to my advisor, Samuel J. Newcome, M. Sc., for suggesting directions and providing valueable feedback.
+
+Computational resources were provided by the e-INFRA CZ project (ID:90254),
+supported by the Ministry of Education, Youth and Sports of the Czech Republic.
 
@@
 However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
+{\color{red} TODO: source}
 \end{itemize}
@@
 \caption{\color{red} TODO}
-\label{fig}
+\label{energy}
 \end{figure}
+
+\begin{table*}[htbp]
+\caption{Comparison of running times taken by different configurations of the simulator}
+\begin{center}
+\begin{tabular}{ccrrrrrrrrcc}
+\toprule
+Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ $^{\mathrm{a}}$ & Iters $^{\mathrm{b}}$ & Abs.
+error $^{\mathrm{c}}$ & Rel.
+error $^{\mathrm{d}}$ & CPU time & Runtime \\
+\midrule
+parareal & $10^4\text{days}$ & 32 & 32 &  5000 &  50000 & 1600000 & 1 & 0.0030 AU & 1.00\% & 00:05:33 & 00:00:35 \\
+parareal & $10^4\text{days}$ & 16 & 16 & 10000 & 100000 & 1600000 & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
+serial & $10^4\text{days}$ & 1 & 1 & N/A & 1600000 & 1600000 & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
+\bottomrule
+\multicolumn{11}{l}{$^{\mathrm{a}}$ Resulting number of steps (equal to number of segments times $N_\text{Fine}$)} \\
+\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}$)} \\
+\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
+\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
+\end{tabular}
+\label{runtimes}
+\end{center}
+\end{table*}
 

### 2024-11-30 paper/main.tex
Added results

@@
 The integration part is the part which usually is dependent on some previous time step and thus can't be easily parallelized.
-In order to solve this problem, multiple techniques were devised with the most widely known one being the parareal algorithm.
+In order to make it parallelizable, multiple techniques were devised, one of them being the parareal algorithm.
 
@@
 \midrule
-parareal & $10^4\text{days}$ & 32 & 32 &  5000 &  50000 & 1600000 & 1 & 0.0030 AU & 1.00\% & 00:05:33 & 00:00:35 \\
+parareal & $10^4\text{days}$ & 32 & 32 &  5000 &  50000 & 1600000 & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
 parareal & $10^4\text{days}$ & 16 & 16 & 10000 & 100000 & 1600000 & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\

### 2024-11-30 paper/fig_3.py

@@ -8,7 +8,7 @@ import ref.planets as planets
 
 
 def fig_3():
-	data = np.loadtxt("parareal_test_out.txt.7069723")
+	data = np.loadtxt("results/32.out.txt")
 	data, energy = data[:, :-2], data[:, -2:]
 
 	data = data.reshape((data.shape[0], data.shape[1]//3, 3))

### 2024-11-30 paper/fig_4.py

@@ -11,7 +11,7 @@ import ref.planets as planets
 
 
 def fig_4():
-	data = np.loadtxt("parareal_test_out.txt.7069723")
+	data = np.loadtxt("results/32.out.txt")
 	data, energy = data[:, :-2], data[:, -2:]
 
 	ax = plt.axes()

### 2024-11-30 parareal/dist.py

@@ -1,37 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-import sys
-
-
-if len(sys.argv) <= 2:
-	print("not enough arguments", file=sys.stderr)
-	print("python3 dist.py <parareal out> <serial out>")
-	sys.exit(1)
-
-data_1 = np.loadtxt(sys.argv[1])
-data_1, energy_1 = data_1[:, :-2], data_1[:, -2:]
-data_1 = data_1.reshape((data_1.shape[0], data_1.shape[1]//3, 3))
-
-data_2 = np.loadtxt(sys.argv[2])
-data_2, energy_2 = data_2[:, :-2], data_2[:, -2:]
-data_2 = data_2.reshape((data_2.shape[0], data_2.shape[1]//3, 3))
-
-data_1 = data_1[:min(len(data_1), len(data_2))]
-data_2 = data_2[:min(len(data_1), len(data_2))]
-
-abs_err = np.linalg.norm(data_1-data_2, axis=-1)
-
-print("max abs err: ", abs_err.max())
-print("mean abs err: ", abs_err.mean())
-
-tolerance = 0.1
-rel_errs = abs_err/np.linalg.norm(data_2, axis=-1)
-print("max rel err: ", rel_errs.max())
-print("mean rel err: ", rel_errs.mean())
-
-print("no. of outliers (rel err > 0.1): ", (rel_errs > tolerance).sum() / len(data_1))
-
-plt.hist(rel_errs[:, 1])
-
-#plt.plot(np.arange(len(data_1)), rel_errs[:, 1])
-plt.show()

### 2024-11-30 scripts/dist.py

@@ -1,0 +1,40 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import sys
+
+
+def compare(parareal_outfile, serial_outfile):
+	data_1 = np.loadtxt(sys.argv[1])
+	data_1, energy_1 = data_1[:, :-2], data_1[:, -2:]
+	data_1 = data_1.reshape((data_1.shape[0], data_1.shape[1]//3, 3))
+
+	data_2 = np.loadtxt(sys.argv[2])
+	data_2, energy_2 = data_2[:, :-2], data_2[:, -2:]
+	data_2 = data_2.reshape((data_2.shape[0], data_2.shape[1]//3, 3))
+
+	data_1 = data_1[:min(len(data_1), len(data_2))]
+	data_2 = data_2[:min(len(data_1), len(data_2))]
+
+	abs_errs = np.linalg.norm(data_1-data_2, axis=-1)
+	rel_errs = abs_errs/np.linalg.norm(data_2, axis=-1)
+
+	return abs_errs, rel_errs
+
+
+if __name__ == "__main__":
+	if len(sys.argv) <= 2:
+		print("not enough arguments", file=sys.stderr)
+		print("python3 dist.py <parareal out> <serial out>")
+		sys.exit(1)
+
+	abs_errs, rel_errs = compare(sys.argv[1], sys.argv[2])
+
+	print(f"max abs err: {abs_errs.max():.4f} AU")
+	print(f"mean abs err: {abs_errs.mean():.4f} AU")
+
+	print(f"max rel err: {rel_errs.max() * 100:.2f} %")
+	print(f"mean rel err: {rel_errs.mean() * 100:.2f} %")
+
+	#plt.hist(rel_errs[:, 1])
+	#plt.plot(np.arange(len(data_1)), rel_errs[:, 1])
+	#plt.show()

### 2024-12-01 paper/main.tex
Improved figures

@@
 \centerline{\includegraphics[width=\linewidth]{fig_1.eps}}
-\caption{Parareal convergence for the example of a harmonic oscillator. The coarse solver is Euler and the fine solver is RK4}
+\caption{Parareal convergence for the example of a harmonic oscillator with $k = 1\frac{\text{N}}{\text{m}}$. The coarse solver is Euler and the fine solver is RK4.}
 \label{oscillator}
@@
 \begin{itemize}
-    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its better precision.
+    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties {\color{red} TODO: source} and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision {\color{red} TODO: source}.
     \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic.
@@
 However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
-{\color{red} TODO: source}
 \end{itemize}
@@
 \centerline{\includegraphics[width=\linewidth]{fig_2.eps}}
-\caption{Path of Mercury over the time of one Earth year in the simulation converging over multiple iterations of the parareal algorithm.
-Velocity verlet and RK4 are used with the same step size of $0.5\text{day}$.}
+\caption{Path of Mercury over the time of one Earth year in a parareal integrated simulation converging over multiple iterations of the parareal algorithm.
+Velocity verlet and RK4 are used with the same step size of $0.5\text{ day}$.}
 \label{mercury}
@@
 \centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
-\caption{\color{red} TODO}
-\label{longrunning}
+\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see table \ref{runtimes}.}
+\label{sim}
 \end{figure}
@@
 \centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
-\caption{\color{red} TODO}
+\caption{Conservation of energy. Same simulation as in Fig. \ref{sim}.}
 \label{energy}

### 2024-12-01 paper/fig_4.py

@@ -20,7 +20,7 @@ def fig_4():
 	ax.plot(np.arange(len(energy))*100, energy.sum(axis=1), label="total energy")
 	ax.legend()
 	plt.xlabel("Time [day]")
-	plt.ylabel(r"Energy [$\textrm{M}_{\textrm{Earth}}\frac{\textrm{AU}^2}{\textrm{day}^2}$]")
+	plt.ylabel(r"Energy $\left[\textrm{M}_{\textrm{Earth}}\frac{\textrm{AU}^2}{\textrm{day}^2}\right]$")
 
 	plt.tight_layout()
 	plt.savefig("fig_4.eps", format="eps")

### 2024-12-01 ref/parareal.py

@@ -140,6 +140,9 @@ def test_problem(show=True):
 
 	plt.plot(ts, np.sin(ts), label=f"true function")
 
+	plt.xlabel("t [s]")
+	plt.ylabel("x [m]")
+
 	plt.legend()
 	if show:
 		plt.show()

### 2024-12-01 paper/main.tex
Paper

@@
 \begin{IEEEkeywords}
-parareal, particle-simulations, parallel-in-time.
+parareal, particle simulations, parallel-in-time, solar system.
 \end{IEEEkeywords}
-
-\section*{Acknowledgment}
-\addcontentsline{toc}{section}{Acknowledgment}
-
-I am greatful to my advisor, Samuel J. Newcome, M. Sc., for suggesting directions and providing valueable feedback.
-
-Computational resources were provided by the e-INFRA CZ project (ID:90254),
-supported by the Ministry of Education, Youth and Sports of the Czech Republic.
 
@@
 We use a simple $n$-body particle simulation of the solar system with time complexity $\mathcal{O}(n^2)$, which we then integrate using the parareal method.
-Thus, we are integrating the following function:
+This is inefficient for a large number of particles, however, our focus is on the parallel-in-time integration, therefore we use a simpler simulation.
+
+Thus, we are integrating the following equation:
 $$
-f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
+\diff{\mathbf{u}(t)}{t} = f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
 $$
@@
 where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from the Newton's law of universal gravitation.
+
+It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver.
+This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.\cite{parallelsolar}\cite{farfuture}
 
@@
 
-\section{Implementation}
+\subsection{Stopping criteria}
+We assume the algorithm has converged when the maximum distance between the states of two consecutive iterations is smaller than some $\varepsilon$. We use $\varepsilon = 10^{-3}\text{ AU}$.
 
@@
 
+\section{Implementation and experiments}
+Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP.
+
+In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI.
+However, due to the higher complexity and difficulty of such an implementation, an easier approach using OpenMP was used in this project for the present time instead.
+
+Multiple experiments were ran on a cluster in order to get a more reproducible environment.
+The resulting run times were collected in Table \ref{runtimes}.
+For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.
+
 \begin{figure}[htbp]
 \centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
-\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see table \ref{runtimes}.}
+\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see Table \ref{runtimes}.}
 \label{sim}
@@
 \begin{table*}[htbp]
-\caption{Comparison of running times taken by different configurations of the simulator}
+\caption{Comparison of running times taken by different configurations of the simulator. 
+$N_{\text{Coarse}}$ and $N_{\text{Fine}}$ refer to the number of steps of the coarse and the fine integrators respectively. 
+To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
+\newline The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
 \begin{center}
@@
 \toprule
-Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ $^{\mathrm{a}}$ & Iters $^{\mathrm{b}}$ & Abs.
-error $^{\mathrm{c}}$ & Rel.
-error $^{\mathrm{d}}$ & CPU time & Runtime \\
+Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs.
+error $^{\mathrm{b}}$ & Rel.
+error $^{\mathrm{c}}$ & CPU time & Runtime \\
 \midrule
-parareal & $10^4\text{days}$ & 32 & 32 &  5000 &  50000 & 1600000 & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
-parareal & $10^4\text{days}$ & 16 & 16 & 10000 & 100000 & 1600000 & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
-serial & $10^4\text{days}$ & 1 & 1 & N/A & 1600000 & 1600000 & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
+parareal & $10^4\text{days}$ & 32 & 32 &  $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
+parareal & $10^4\text{days}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
+serial & $10^4\text{days}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
 \bottomrule
-\multicolumn{11}{l}{$^{\mathrm{a}}$ Resulting number of steps (equal to number of segments times $N_\text{Fine}$)} \\
-\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}$)} \\
+\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
 \multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
-\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
+\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation} \\
 \end{tabular}
@@
 \end{table*}
+
+\section{Summary and discussion}
+
+\section*{Acknowledgment}
+\addcontentsline{toc}{section}{Acknowledgment}
+
+I would like to thank my advisor, Samuel J. Newcome, M. Sc., for suggesting directions and providing valueable feedback.
+
+Computational resources were provided by the e-INFRA CZ project (ID:90254),
+supported by the Ministry of Education, Youth and Sports of the Czech Republic.
 
@@
 409 (1997,7), http://dx.doi.org/10.1086/118485
+\bibitem{farfuture}Gander, M.
+Is it possible to predict the far future before the near future is known accurately?.
+{\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}.
+pp.
+21 (2019), [Online].
+Available: http://publications.mfo.de/handle/mfo/3693
 \bibitem{parareal2}Gander, M.

### 2024-12-01 paper/fig_3.py

@@ -8,7 +8,7 @@ import ref.planets as planets
 
 
 def fig_3():
-	data = np.loadtxt("results/32.out.txt")
+	data = np.loadtxt("results/m.32.out.txt")
 	data, energy = data[:, :-2], data[:, -2:]
 
 	data = data.reshape((data.shape[0], data.shape[1]//3, 3))

### 2024-12-01 paper/fig_4.py

@@ -11,7 +11,7 @@ import ref.planets as planets
 
 
 def fig_4():
-	data = np.loadtxt("results/32.out.txt")
+	data = np.loadtxt("results/m.32.out.txt")
 	data, energy = data[:, :-2], data[:, -2:]
 
 	ax = plt.axes()

### 2024-12-01 parareal/job.sh
New job

@@ -1,7 +1,7 @@
 #!/bin/bash
 #PBS -N parareal_test
-#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb
-#PBS -l walltime=00:10:00
+#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zenon
+#PBS -l walltime=01:30:00
 
 export OMP_NUM_THREADS=$PBS_NUM_PPN
 
@@ -13,15 +13,14 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 echo "$PBS_JOBID is running on node `hostname -f`" >> $DATADIR/jobs_info.txt
 
 # run program
-NSEG=32
-COARSE_STEPS=5000
-FINE_STEPS=50000
-TIME=10000
-EPS=1e-3
+NSEG=64
+COARSE_STEPS=100000
+FINE_STEPS=1000000
+TIME=100000
+EPS=1e-4
 MAX_ITERS=16
-LOG_PERIOD=10
+LOG_PERIOD=1000
 
-echo "=== parareal ==="
 time $DATADIR/build/parareal -i $DATADIR/planets.cfg -o $DATADIR/out_p.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
 
 #echo "===  serial  ==="

### 2024-12-01 parareal/test.sh

@@ -3,12 +3,12 @@ cd build
 cmake --build .
 
 NSEG=32
-COARSE_STEPS=5000
-FINE_STEPS=50000
-TIME=10000
+COARSE_STEPS=200000
+FINE_STEPS=2000000
+TIME=100000
 EPS=1e-3
-MAX_ITERS=8
-LOG_PERIOD=10
+MAX_ITERS=16
+LOG_PERIOD=1000
 
 echo "=== parareal ==="
 time ./parareal -i ../planets.cfg -o out_p.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD

### 2024-12-02 paper/main.tex
Fixed bugs

@@
 \usepackage{xcolor}
+\usepackage{comment}
 \usepackage[thinc]{esdiff}
@@
 This is repeated until satisfactory convergence.
-It has been proven for example by \cite{parareal} (See Proposition 1) that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
+It has been proven for example by \cite{parareal} (See Proposition 1) {\color{red} TODO: better source} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
 
@@
 \end{figure}
+
+\subsection*{Stopping criteria}
+We assume the algorithm has converged enough when $\forall n: |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist.\cite{speedup}
+
+\subsection*{Theoretical speed-up}
+
+A derivation of the theoretical speedup is presented in \cite{speedup} (Section 2.3), which we use here.
+It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially.
+
+To find the theoretical speed-up of the algorithm, let's compute the time required by the parareal algorithm, $T_\text{parareal}$, and the time $T_\text{serial}$ taken to compute the solution by running the fine integrator serially.
+
+Assume running $\mathcal{F}$ over any segment $[t_n, t_{n+1}]$ takes $T_\mathcal{F}$ and running $\mathcal{C}$ over any segment takes $T_\mathcal{C}$. Then computing $\mathcal{F}$ serially takes
+$$
+T_\text{serial} \approx NT_\mathcal{F}
+$$
+while using the parareal algorithm, with $N$ CPUs, we have
+$$
+T_\text{parareal} \approx NT_\mathcal{C} + \sum_{i=1}^{k}\left(T_\mathcal{F} + NT_\mathcal{C}\right) = kT_\mathcal{F} + (k+1)NT_\mathcal{C}
+$$
+
+Finally, the speed-up $S_\text{parareal}$ can be computed as
+\begin{multline}
+S_\text{parareal} = \frac{T_\text{serial}}{T_\text{parareal}} = \\\frac{NT_\mathcal{F}}{kT_\mathcal{F} + (k+1)NT_\mathcal{C}} = \\\left(\frac{k}{N} + (k+1)\frac{T_\mathcal{C}}{T_\mathcal{F}}\right)^{-1}
+\end{multline}
+
+Thus, to reach a high speed-up, we want the ratio $\frac{T_\mathcal{C}}{T_\mathcal{F}}$ to be small while also keeping the number of required iterations $k$ small.\cite{speedup}
+
+The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up.\cite{speedup}
 
@@
 \begin{itemize}
-    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties {\color{red} TODO: source} and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision {\color{red} TODO: source}.
+    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties {\color{red} TODO: source} and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision.
     \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic.
@@
 \end{itemize}
-
-\subsection{Stopping criteria}
-We assume the algorithm has converged when the maximum distance between the states of two consecutive iterations is smaller than some $\varepsilon$. We use $\varepsilon = 10^{-3}\text{ AU}$.
 
@@
 \section{Implementation and experiments}
-Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP.
+Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP. {\color{red} TODO: rewrite}
 
@@
 For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.
+
+Let's derive the expected speed-up of our program and compare it with the measured speed-up.
+To simplify the calculation, we will only estimate it using the number of function evaluations.
+For Velocity Verlet, we have $T_\mathcal{C} = 1N_\mathcal{C}$, where $N_\mathcal{C}$ is the number of steps of the coarse integrator.
+RK4 takes four function evaluations, so we have $T_\mathcal{F} = 4N_\mathcal{F}$.
+If the fine integrator does 10 times more steps than the coarse one, we have $\frac{N_\mathcal{C}}{N_\mathcal{F}} = \frac{1}{10}$.
+Thus, we have
+$$
+S_\text{parareal} = \left(\frac{k}{N} + (k+1)\frac{1N_\mathcal{C}}{4N_\mathcal{F}}\right)^{-1} = \left(\frac{k}{N} + \frac{k+1}{40}\right)^{-1}
+$$
+When we choose the number of processors (and the number of segments), $N = 16$ and we converge within one iteration, we have
+$$
+S_\text{parareal} = \left(\frac{1}{16} + \frac{1}{20}\right)^{-1} \approx 8.89
+$$
 
@@
 
+\begin{comment}
 \begin{table*}[htbp]
@@
 $N_{\text{Coarse}}$ and $N_{\text{Fine}}$ refer to the number of steps of the coarse and the fine integrators respectively. 
-To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
-\newline The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
+To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).}
 \begin{center}
@@
 \toprule
-Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs.
-error $^{\mathrm{b}}$ & Rel.
-error $^{\mathrm{c}}$ & CPU time & Runtime \\
+Program & Environment & n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs.
+err.
+$^{\mathrm{b}}$ & Rel.
+err.
+$^{\mathrm{c}}$ & CPU time & Runtime \\
 \midrule
-parareal & $10^4\text{days}$ & 32 & 32 &  $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
-parareal & $10^4\text{days}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
-serial & $10^4\text{days}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
+\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
+\midrule[.025em]
+parareal & zia.cerit-sc.cz$^{\mathrm{d}}$ & 32 & 32 & $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
+parareal & zia.cerit-sc.cz$^{\mathrm{d}}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
+serial & zia.cerit-sc.cz$^{\mathrm{d}}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
 \bottomrule
-\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
-\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
-\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation} \\
+\multicolumn{11}{l}{$^{\mathrm{a}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
+\multicolumn{11}{l}{$^{\mathrm{b}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
+\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum relative error across the trajectory compared to the serial computation} \\
+\multicolumn{11}{l}{$^{\mathrm{d}}$ The programs were run on the machines of the zia.cerit-sc.cz cluster } \\
+\multicolumn{11}{l}{owned by CERIT-SC/MU, a part of the MetaCentrum organization,} \\
+\multicolumn{11}{l}{each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
 \end{tabular}
@@
 \end{table*}
+\end{comment}
 
@@
 409 (1997,7), http://dx.doi.org/10.1086/118485
+\bibitem{speedup}Pentland, K., Tamborrino, M., Sullivan, T., Buchanan, J.
+\& Appel, L.
+GParareal: a time-parallel ODE solver using Gaussian process emulation.
+{\em Statistics And Computing}.
+33, 23 (2022,12), [Online].
+Available: https://doi.org/10.1007/s11222-022-10195-y
 \bibitem{farfuture}Gander, M.

### 2024-12-02 parareal/src/integration.hpp

@@ -13,29 +13,37 @@ namespace parareal { namespace integration {
 		void operator()(double t0, double t1, state::State& state) {
 			double dt = t1-t0;
 
+			state.compute_forces();
 			for (std::size_t i = 0; i < state.size(); ++i) {
-				state.compute_force(i);
 				state.x[i] += state.v[i]*dt;
 				state.v[i] += state.a[i]*dt;
 			}
 		}
 	};
 
-	class VerletStep {
+	class VelocityVerlet {
+	private:
+		std::size_t num_steps;
+
 	public:
+		VelocityVerlet(std::size_t num_steps): num_steps(num_steps) {};
+
 		void operator()(double t0, double t1, state::State& state) {
-			double dt = t1-t0;
+			double dt = (t1-t0)/num_steps;
+			state.compute_forces();
 
-			for (std::size_t i = 0; i < state.size(); ++i) {
-				state.compute_force(i);
-				const types::vec3 a0 = state.a[i];
+			for (std::size_t it = 0; it < num_steps; ++it) {
+				for (std::size_t i = 0; i < state.size(); ++i) {
+					state.x[i] += dt * (state.v[i] + 0.5*state.a[i]*dt);
+				}
 
-				state.x[i] += state.v[i]*dt + a0/2.*dt*dt;
-
-				state.compute_force(i);
-				const types::vec3 a1 = state.a[i];
-
-				state.v[i] += (a0+a1)/2. * dt;
+				for (std::size_t i = 0; i < state.size(); ++i) {
+					// assumption: accelerations don't depend on velocities
+					const types::vec3 a0 = state.a[i];
+					state.compute_force(i);
+					const types::vec3 a1 = state.a[i];
+					state.v[i] += (a0+a1)/2. * dt;
+				}
 			}
 		}
 	};
@@ -45,46 +53,55 @@ namespace parareal { namespace integration {
 		void operator()(double t0, double t1, state::State& state) {
 			double dt = t1-t0;
 
-			types::vec3 k1_v;
-			types::vec3 k2_v;
-			types::vec3 k3_v;
-			types::vec3 k4_v;
-			types::vec3 k1_x;
-			types::vec3 k2_x;
[…]

### 2024-12-02 parareal/src/main.cpp

@@ -51,7 +51,7 @@ int main(int argc, char** argv) {
 		if (args["serial"].as<bool>()) {
 			auto steps = nseg*fsteps;
 
-			auto solver = parareal::integration::solver_with_log_period<parareal::integration::RK4Step>(steps, log_period);
+			auto solver = parareal::integration::solver_with_log_period<parareal::integration::RK4Fixed>(steps, log_period);
 			res = solver.integrate_verbose(0, time, initial);
 		} else {
 			res = parareal::integration::parareal(
@@ -61,8 +61,8 @@ int main(int argc, char** argv) {
 				nseg,                                // number of segments (must be divisible by the number of ranks)
 				args["max-iters"].as<std::size_t>(), // maximum of parareal iters, 
 				args["eps"].as<double>(),            // eps for convergence
-				parareal::integration::FixedStepSolver<parareal::integration::VerletStep>(csteps),
-				parareal::integration::solver_with_log_period<parareal::integration::RK4Step>(fsteps, log_period)
+				parareal::integration::VelocityVerlet(csteps),
+				parareal::integration::solver_with_log_period<parareal::integration::RK4Fixed>(fsteps, log_period)
 			);
 		}
 

### 2024-12-02 parareal/src/state.hpp

@@ -45,6 +45,12 @@ namespace parareal { namespace state {
 			return x.size();
 		}
 
+		void compute_forces() {
+			for (std::size_t i = 0; i < size(); ++i) {
+				compute_force(i);
+			}
+		}
+
 		void compute_force(std::size_t i) {
 			a[i].clear();
 

### 2024-12-02 parareal/test.sh

@@ -2,20 +2,17 @@
 cd build
 cmake --build .
 
-NSEG=32
-COARSE_STEPS=200000
-FINE_STEPS=2000000
-TIME=100000
-EPS=1e-3
-MAX_ITERS=16
-LOG_PERIOD=1000
+NSEG=16
+COARSE_STEPS=1000
+FINE_STEPS=10000
+TIME=10000
+EPS=1e-4
+MAX_ITERS=8
+LOG_PERIOD=10
 
 echo "=== parareal ==="
 time ./parareal -i ../planets.cfg -o out_p.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
-echo "===  serial  ==="
-time ./parareal --serial -i ../planets.cfg -o out.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
-
-echo "=== distance ==="
-python3 ../dist.py out_p.txt out.txt
+#echo "===  serial  ==="
+#time ./parareal --serial -i ../planets.cfg -o out.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
 
 #python3 vis.py out_p.txt

### 2024-12-02 scripts/dist.py

@@ -35,6 +35,10 @@ if __name__ == "__main__":
 	print(f"max rel err: {rel_errs.max() * 100:.2f} %")
 	print(f"mean rel err: {rel_errs.mean() * 100:.2f} %")
 
-	#plt.hist(rel_errs[:, 1])
+	print(f"ratio of outliers (rel. err. > 0.3): {(rel_errs > 0.3).sum()/len(rel_errs)}")
+
+	for i in range(rel_errs.shape[1]):
+		plt.hist(rel_errs[:, i], bins=100, label=f"{i}")
+	plt.legend()
 	#plt.plot(np.arange(len(data_1)), rel_errs[:, 1])
-	#plt.show()
+	plt.show()

### 2024-12-02 paper/main.tex
Speed-up

@@
 
-Let's derive the expected speed-up of our program and compare it with the measured speed-up.
-To simplify the calculation, we will only estimate it using the number of function evaluations.
-For Velocity Verlet, we have $T_\mathcal{C} = 1N_\mathcal{C}$, where $N_\mathcal{C}$ is the number of steps of the coarse integrator.
-RK4 takes four function evaluations, so we have $T_\mathcal{F} = 4N_\mathcal{F}$.
-If the fine integrator does 10 times more steps than the coarse one, we have $\frac{N_\mathcal{C}}{N_\mathcal{F}} = \frac{1}{10}$.
-Thus, we have
-$$
-S_\text{parareal} = \left(\frac{k}{N} + (k+1)\frac{1N_\mathcal{C}}{4N_\mathcal{F}}\right)^{-1} = \left(\frac{k}{N} + \frac{k+1}{40}\right)^{-1}
-$$
-When we choose the number of processors (and the number of segments), $N = 16$ and we converge within one iteration, we have
-$$
-S_\text{parareal} = \left(\frac{1}{16} + \frac{1}{20}\right)^{-1} \approx 8.89
-$$
-
 \begin{figure}[htbp]

### 2024-12-02 parareal/job.sh
Better job structure

@@ -1,32 +1,0 @@
-#!/bin/bash
-#PBS -N parareal_test
-#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zenon
-#PBS -l walltime=01:30:00
-
-export OMP_NUM_THREADS=$PBS_NUM_PPN
-
-# define a DATADIR variable: directory where the input files are taken from and where the output will be copied to
-DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
-
-# append a line to a file "jobs_info.txt" containing the ID of the job, the hostname of the node it is run on, and the path to a scratch directory
-# this information helps to find a scratch directory in case the job fails, and you need to remove the scratch directory manually 
-echo "$PBS_JOBID is running on node `hostname -f`" >> $DATADIR/jobs_info.txt
-
-# run program
-NSEG=64
-COARSE_STEPS=100000
-FINE_STEPS=1000000
-TIME=100000
-EPS=1e-4
-MAX_ITERS=16
-LOG_PERIOD=1000
-
-time $DATADIR/build/parareal -i $DATADIR/planets.cfg -o $DATADIR/out_p.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
-
-#echo "===  serial  ==="
-#time ./parareal --serial -i $DATADIR/giants.cfg -o out.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS
-#echo "=== distance ==="
-#python3 $DATADIR/dist.py out_p.txt out.txt
-
-# clean the SCRATCH directory
-# clean_scratch

### 2024-12-02 parareal/jobs/32.10^5.cfg.sh

@@ -1,0 +1,7 @@
+NSEG=32
+MAX_ITERS=16
+COARSE_STEPS=10000
+FINE_STEPS=100000
+TIME=10000
+EPS=1e-3
+LOG_PERIOD=100

### 2024-12-02 parareal/jobs/base_job.sh

@@ -1,0 +1,14 @@
+export OMP_NUM_THREADS=$PBS_NUM_PPN
+
+# define a DATADIR variable: directory where the input files are taken from and where the output will be copied to
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
+echo time $DATADIR/build/parareal -i $DATADIR/planets.cfg -o $DATADIR/results/$OUT --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD $FLAGS
+
+mkdir -p $DATADIR/results/
+# append a line to a file "jobs_info.txt" containing the ID of the job, the hostname of the node it is run on, and the path to a scratch directory
+# this information helps to find a scratch directory in case the job fails, and you need to remove the scratch directory manually 
+echo "$PBS_JOBID is running on node `hostname -f`" >> $DATADIR/results/jobs_info.txt
+
+# run program
+time $DATADIR/build/parareal -i $DATADIR/planets.cfg -o $DATADIR/results/$OUT --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD $FLAGS

### 2024-12-02 parareal/jobs/job.32.32.zenon.10^5.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+#PBS -N parareal_32_32_test
+#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zenon
+#PBS -l walltime=00:05:00
+
+# config
+source 32.10^5.cfg.sh
+
+# run
+OUT="out.32.32.zenon.10^5.txt"
+FLAGS=""
+
+source base_job.sh

### 2024-12-02 parareal/jobs/job.serial.32.zenon.10^5.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+#PBS -N parareal_1_32_test
+#PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zenon
+#PBS -l walltime=00:05:00
+
+# config
+source 32.10^5.cfg.sh
+
+# run
+OUT="out.serial.32.zenon.10^5.txt"
+FLAGS="--serial"
+
+source base_job.sh

### 2024-12-02 parareal/src/integration.hpp

@@ -42,7 +42,7 @@ namespace parareal { namespace integration {
 					const types::vec3 a0 = state.a[i];
 					state.compute_force(i);
 					const types::vec3 a1 = state.a[i];
-					state.v[i] += (a0+a1)/2. * dt;
+					state.v[i] += 0.5*dt*a0 + 0.5*dt*a1;
 				}
 			}
 		}

### 2024-12-02 parareal/src/state.hpp

@@ -7,7 +7,7 @@
 
 #include "types.hpp"
 
-#define SMOOTHING_EPS 0
+//#define SMOOTHING_EPS 0
 
 
 namespace parareal { namespace state {
@@ -62,9 +62,9 @@ namespace parareal { namespace state {
 				auto diff = x[i]-x[j];
 				auto dist = diff.norm();
 
-				auto smoothed = std::sqrt(dist*dist + SMOOTHING_EPS*SMOOTHING_EPS);
+				//auto smoothed = std::sqrt(dist*dist + SMOOTHING_EPS*SMOOTHING_EPS);
 
-				a[i] += -cfg->G * cfg->m[j] * diff / std::pow(smoothed, 3.);
+				a[i] += -cfg->G * cfg->m[j] * diff / std::pow(dist, 3.);
 			}
 		}
 
@@ -80,9 +80,9 @@ namespace parareal { namespace state {
 					auto diff = x[i]-x[j];
 					auto dist = diff.norm();
 
-					auto smoothed = std::sqrt(dist*dist + SMOOTHING_EPS*SMOOTHING_EPS);
+					//auto smoothed = std::sqrt(dist*dist + SMOOTHING_EPS*SMOOTHING_EPS);
 
-					energy += -cfg->G * cfg->m[i] * cfg->m[j] / smoothed / 2.;
+					energy += -cfg->G * cfg->m[i] * cfg->m[j] / dist / 2.;
 				}
 			}
 

### 2024-12-02 parareal/test.sh

@@ -1,18 +1,0 @@
-#!/bin/bash
-cd build
-cmake --build .
-
-NSEG=16
-COARSE_STEPS=1000
-FINE_STEPS=10000
-TIME=10000
-EPS=1e-4
-MAX_ITERS=8
-LOG_PERIOD=10
-
-echo "=== parareal ==="
-time ./parareal -i ../planets.cfg -o out_p.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
-#echo "===  serial  ==="
-#time ./parareal --serial -i ../planets.cfg -o out.txt --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD
-
-#python3 vis.py out_p.txt

### 2024-12-02 parareal/jobs/base_job.sh
Removed echo from job

@@ -3,8 +3,6 @@ export OMP_NUM_THREADS=$PBS_NUM_PPN
 # define a DATADIR variable: directory where the input files are taken from and where the output will be copied to
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
-echo time $DATADIR/build/parareal -i $DATADIR/planets.cfg -o $DATADIR/results/$OUT --nseg $NSEG --coarse-steps $COARSE_STEPS --fine-steps $FINE_STEPS --time $TIME --max-iters $MAX_ITERS --eps $EPS --log-period $LOG_PERIOD $FLAGS
-
 mkdir -p $DATADIR/results/
 # append a line to a file "jobs_info.txt" containing the ID of the job, the hostname of the node it is run on, and the path to a scratch directory
 # this information helps to find a scratch directory in case the job fails, and you need to remove the scratch directory manually 

### 2024-12-02 parareal/jobs/base_job.sh
Fixed jobs

@@ -1,8 +1,5 @@
 export OMP_NUM_THREADS=$PBS_NUM_PPN
 
-# define a DATADIR variable: directory where the input files are taken from and where the output will be copied to
-DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
-
 mkdir -p $DATADIR/results/
 # append a line to a file "jobs_info.txt" containing the ID of the job, the hostname of the node it is run on, and the path to a scratch directory
 # this information helps to find a scratch directory in case the job fails, and you need to remove the scratch directory manually 

### 2024-12-02 parareal/jobs/job.32.32.zenon.10^5.sh

@@ -3,11 +3,13 @@
 #PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zenon
 #PBS -l walltime=00:05:00
 
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
 # config
-source 32.10^5.cfg.sh
+source $DATADIR/jobs/32.10^5.cfg.sh
 
 # run
 OUT="out.32.32.zenon.10^5.txt"
 FLAGS=""
 
-source base_job.sh
+source $DATADIR/jobs/base_job.sh

### 2024-12-02 parareal/jobs/job.serial.32.zenon.10^5.sh

@@ -3,11 +3,13 @@
 #PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zenon
 #PBS -l walltime=00:05:00
 
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
 # config
-source 32.10^5.cfg.sh
+source $DATADIR/jobs/32.10^5.cfg.sh
 
 # run
 OUT="out.serial.32.zenon.10^5.txt"
 FLAGS="--serial"
 
-source base_job.sh
+source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.32.32.zia.10^4.sh
Changed cluster; Renamed

@@ -1,6 +1,6 @@
 #!/bin/bash
-#PBS -N parareal_1_32_test
-#PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zenon
+#PBS -N parareal_32_32_test
+#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zia
 #PBS -l walltime=00:05:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
@@ -9,7 +9,7 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 source $DATADIR/jobs/32.10^5.cfg.sh
 
 # run
-OUT="out.serial.32.zenon.10^5.txt"
-FLAGS="--serial"
+OUT="out.32.32.zia.10^5.txt"
+FLAGS=""
 
 source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.serial.32.zia.10^4.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
-#PBS -N parareal_32_32_test
-#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zenon
+#PBS -N parareal_1_32_test
+#PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zia
 #PBS -l walltime=00:05:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
@@ -9,7 +9,7 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 source $DATADIR/jobs/32.10^5.cfg.sh
 
 # run
-OUT="out.32.32.zenon.10^5.txt"
-FLAGS=""
+OUT="out.serial.32.zia.10^4.txt"
+FLAGS="--serial"
 
 source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.32.32.zia.10^4.sh
Fixed jobs

@@ -1,15 +1,15 @@
 #!/bin/bash
 #PBS -N parareal_32_32_test
 #PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zia
-#PBS -l walltime=00:05:00
+#PBS -l walltime=00:12:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # config
-source $DATADIR/jobs/32.10^5.cfg.sh
+source $DATADIR/jobs/32.10^4.cfg.sh
 
 # run
-OUT="out.32.32.zia.10^5.txt"
+OUT="out.32.32.zia.10^4.txt"
 FLAGS=""
 
 source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.serial.32.zia.10^4.sh

@@ -1,12 +1,12 @@
 #!/bin/bash
 #PBS -N parareal_1_32_test
 #PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zia
-#PBS -l walltime=00:05:00
+#PBS -l walltime=00:15:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # config
-source $DATADIR/jobs/32.10^5.cfg.sh
+source $DATADIR/jobs/32.10^4.cfg.sh
 
 # run
 OUT="out.serial.32.zia.10^4.txt"

### 2024-12-03 parareal/jobs/32.10^4.cfg.sh
Jobs

@@ -1,7 +1,0 @@
-NSEG=32
-MAX_ITERS=16
-COARSE_STEPS=10000
-FINE_STEPS=100000
-TIME=10000
-EPS=1e-3
-LOG_PERIOD=100

### 2024-12-03 parareal/jobs/job.16.16.zia.10^4.sh

@@ -1,21 +1,21 @@
 #!/bin/bash
-#PBS -N parareal_32_32_test
-#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zia
+#PBS -N parareal_64_64_test
+#PBS -l select=1:ncpus=16:ompthreads=16:mem=1gb:cluster=zia
 #PBS -l walltime=00:12:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # config
-NSEG=32
-MAX_ITERS=16
-COARSE_STEPS=10000
-FINE_STEPS=100000
+NSEG=16
+MAX_ITERS=8
+COARSE_STEPS=20000
+FINE_STEPS=200000
 TIME=10000
 EPS=1e-3
 LOG_PERIOD=100
 
 # run
-OUT="out.32.32.zia.10^4.txt"
+OUT="out.16.16.zia.10^4.txt"
 FLAGS=""
 
 source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.64.64.zia.10^4.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+#PBS -N parareal_64_64_test
+#PBS -l select=1:ncpus=64:ompthreads=64:mem=1gb:cluster=zia
+#PBS -l walltime=00:09:00
+
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
+# config
+NSEG=64
+MAX_ITERS=16
+COARSE_STEPS=5000
+FINE_STEPS=50000
+TIME=10000
+EPS=1e-3
+LOG_PERIOD=100
+
+# run
+OUT="out.64.64.zia.10^4.txt"
+FLAGS=""
+
+source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.serial.zia.10^4.sh

@@ -1,15 +1,21 @@
 #!/bin/bash
-#PBS -N parareal_1_32_test
+#PBS -N parareal_serial_test
 #PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zia
 #PBS -l walltime=00:15:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # config
-source $DATADIR/jobs/32.10^4.cfg.sh
+NSEG=32
+MAX_ITERS=16
+COARSE_STEPS=10000
+FINE_STEPS=100000
+TIME=10000
+EPS=1e-3
+LOG_PERIOD=100
 
 # run
-OUT="out.serial.32.zia.10^4.txt"
+OUT="out.serial.zia.10^4.txt"
 FLAGS="--serial"
 
 source $DATADIR/jobs/base_job.sh

### 2024-12-03 parareal/jobs/job.16.16.zia.10^4.sh
Fix job16 name

@@ -1,5 +1,5 @@
 #!/bin/bash
-#PBS -N parareal_64_64_test
+#PBS -N parareal_16_16_test
 #PBS -l select=1:ncpus=16:ompthreads=16:mem=1gb:cluster=zia
 #PBS -l walltime=00:12:00
 

### 2024-12-04 parareal/jobs/job.4.4.zia.10^4.sh
Job 4

@@ -1,0 +1,21 @@
+#!/bin/bash
+#PBS -N parareal_4_4_test
+#PBS -l select=1:ncpus=4:ompthreads=4:mem=1gb:cluster=zia
+#PBS -l walltime=00:10:00
+
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
+# config
+NSEG=4
+MAX_ITERS=1
+COARSE_STEPS=20000
+FINE_STEPS=200000
+TIME=10000
+EPS=1e-3
+LOG_PERIOD=100
+
+# run
+OUT="out.4.4.zia.10^4.txt"
+FLAGS=""
+
+source $DATADIR/jobs/base_job.sh

### 2024-12-04 parareal/jobs/job.4.4.zia.10^4.sh
Job 4

@@ -7,7 +7,7 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # config
 NSEG=4
-MAX_ITERS=1
+MAX_ITERS=2
 COARSE_STEPS=20000
 FINE_STEPS=200000
 TIME=10000

### 2024-12-04 parareal/jobs/job.4.4.zia.10^4.sh
Job 4

@@ -8,8 +8,8 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 # config
 NSEG=4
 MAX_ITERS=2
-COARSE_STEPS=20000
-FINE_STEPS=200000
+COARSE_STEPS=80000
+FINE_STEPS=800000
 TIME=10000
 EPS=1e-3
 LOG_PERIOD=100

### 2024-12-05 parareal/jobs/long.zia.sh
Long sim

@@ -1,0 +1,21 @@
+#!/bin/bash
+#PBS -N parareal_long_test
+#PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zia
+#PBS -l walltime=00:15:00
+
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
+# config
+NSEG=32
+MAX_ITERS=12
+COARSE_STEPS=100000
+FINE_STEPS=1000000
+TIME=365250
+EPS=1e-3
+LOG_PERIOD=1000
+
+# run
+OUT="out.long.zia.txt"
+FLAGS=""
+
+source $DATADIR/jobs/base_job.sh

### 2024-12-05 scripts/verify-pot.py
Moved vis.py; Added script to verify potential energy

@@ -1,0 +1,30 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import sys
+import os
+sys.path.insert(0, os.path.join(os.path.dirname(__file__), "../ref"))
+import planets
+
+
+if __name__ == "__main__":
+	if len(sys.argv) <= 1:
+		print("not enough arguments", file=sys.stderr)
+		sys.exit(1)
+
+	data = np.loadtxt(sys.argv[1])
+	data, energy = data[:, :-2], data[:, -2:]
+	data = data.reshape((data.shape[0], data.shape[1]//3, 3))
+
+	for state, saved in zip(data, energy[:, 1]):
+		correct = 0
+		for i in range(len(state)):
+			for j in range(len(state)):
+				if i == j:
+					continue
+					
+				dist = np.linalg.norm(state[i] - state[j])
+
+				correct += -planets.G * planets.masses[i] * planets.masses[j]/dist/2
+
+		diff = np.abs(correct-saved)
+		print(f"{correct=}, {saved=}, {diff=}")

### 2024-12-05 scripts/vis.py

@@ -1,0 +1,41 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import sys
+
+
+if len(sys.argv) <= 1:
+	print("not enough arguments", file=sys.stderr)
+	sys.exit(1)
+
+data = np.loadtxt(sys.argv[1])
+data, energy = data[:, :-2], data[:, -2:]
+
+data = data.reshape((data.shape[0], data.shape[1]//3, 3))
+
+lim = 1.2*max(data[0, :, :].min(), data[0, :, :].max())
+
+ax = plt.axes(projection="3d")
+ax.set_xlim3d(-lim, lim)
+ax.set_ylim3d(-lim, lim)
+ax.set_zlim3d(-lim, lim)
+
+for i in range(data.shape[1]):
+	ax.plot3D(data[:, i, 0], data[:, i, 1], data[:, i, 2], label=f"particle {i}")
+ax.legend()
+plt.show()
+
+ax = plt.axes()
+ax.plot(np.arange(len(energy)), energy[:, 0], label="kinetic energy")
+ax.plot(np.arange(len(energy)), energy[:, 1], label="potential energy")
+ax.plot(np.arange(len(energy)), energy.sum(axis=1), label="total energy")
+ax.legend()
+plt.show()
+
+ax = plt.axes()
+ax.set_xlim([-lim, lim])
+ax.set_ylim([-lim, lim])
+particles_plot = ax.scatter(data[0, :, 0], data[0, :, 1])
+
+for i in range(len(data)):
+	particles_plot.set_offsets(np.c_[data[i, :, 0], data[i, :, 1]])
+	plt.pause(0.01)

### 2024-12-05 paper/main.tex
Paper: draft

--- 2024-12-02 paper/main.tex
@@
 \usepackage[thinc]{esdiff}
+\usepackage{xurl}
 
@@
 \begin{abstract}
-TBD
+The purpose of this paper is to examine the applications of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
+We then test and measure the real speed-up and compare it with the expectation derived from theory.
 \end{abstract}
@@
 
-Simulating those systems usually requires huge computations, which can take a very long time.
+Simulating those systems usually requires a huge amount of computations, which can take a very long time.
 Some of these computations can be run in parallel in order to reduce the computation time.
-There has been more focus in parallelizing these simulations through the spatial domain, which allows us to run bigger simulations in the same time.
+There has been a lot of focus in parallelizing these simulations through the spatial domain, which allows us to run bigger simulations while taking the same time.
 
-There do, however, exist algorithms that allow us to parallelize simulations in the time domain.
+There do, however, also exist algorithms that allow us to parallelize simulations in the time domain.
 This could be very useful for running smaller simulations through a very long timespan.
@@
 
-A description of the parareal algorithm is given in \cite{parareal}.
-It achieves parallel-in-time integration using two types of integrators.
+The parareal algorithm was introduced by \cite{parareal}.\footnote{The original paper in French, however it includes a description in English.} It achieves parallel-in-time integration using two types of integrators.
 A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to fill in the gaps in this estimated trajectory.
@@
 This is repeated until satisfactory convergence.
-It has been proven for example by \cite{parareal} (See Proposition 1) {\color{red} TODO: better source} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
+It has been proven for example by \cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
 
@@
 A derivation of the theoretical speedup is presented in \cite{speedup} (Section 2.3), which we use here.
-It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially.
+It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially (our implementation doesn't include this optimization).
 
@@
 \begin{itemize}
-    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties {\color{red} TODO: source} and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision.
+    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties with a small cost and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision.
     \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic.
@@
 \section{Implementation and experiments}
-Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP. {\color{red} TODO: rewrite}
 
+\subsection{Implementation}
+An implementation in C++ using OpenMP for the parallelization was written.
 In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI.
@@
 
-Multiple experiments were ran on a cluster in order to get a more reproducible environment.
-The resulting run times were collected in Table \ref{runtimes}.
-For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.
+\subsection{Experiments (for run time comparison)}
+Multiple experiments with the same initial conditions and same step size were ran on a cluster in order to get a more reproducible and comparable environment.
 
-Let's derive the expected speed-up of our program and compare it with the measured speed-up.
+A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration).
+
+The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.
+
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
+\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see Table \ref{runtimes}.}
+\label{sim}
+\end{figure}
+
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
+\caption{Conservation of energy in the simulation. Same simulation as in Fig. \ref{sim}.}
+\label{energy}
+\end{figure}
+
+\subsection{Run time comparison}
+Let's derive the expected speed-up and run time of our program and compare it with the measured run time.
 To simplify the calculation, we will only estimate it using the number of function evaluations.
@@
 
+We can compare this with the real measured speed-up, see Table \ref{runtimes}.
+We observe that in reality the program runs around 5 times faster with 16 threads.
+We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure \ref{speeds}.
+
 \begin{figure}[htbp]
-\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
-\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see Table \ref{runtimes}.}
-\label{sim}
+\centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
+\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPUs (blue points).}
+\label{speeds}
 \end{figure}
 
-\begin{figure}[htbp]
-\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
-\caption{Conservation of energy. Same simulation as in Fig. \ref{sim}.}
-\label{energy}
-\end{figure}
-
-\begin{comment}
 \begin{table*}[htbp]
 \caption{Comparison of running times taken by different configurations of the simulator. 
-$N_{\text{Coarse}}$ and $N_{\text{Fine}}$ refer to the number of steps of the coarse and the fine integrators respectively. 
-To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).}
+$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively. 
+To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
+The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
 \begin{center}
-\begin{tabular}{ccrrrrrrrrcc}
+\begin{tabular}{crrrrrrrrrcc}
 \toprule
-Program & Environment & n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs.
+Program & Time & n CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_{\text{Final}}$ & Iters $^{\mathrm{b}}$ & Abs.
 err.
-$^{\mathrm{b}}$ & Rel.
+$^{\mathrm{c}}$ & Rel.
 err.
-$^{\mathrm{c}}$ & CPU time & Runtime \\
+$^{\mathrm{d}}$ & CPU time & Real time \\
 \midrule
-\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
-\midrule[.025em]
-parareal & zia.cerit-sc.cz$^{\mathrm{d}}$ & 32 & 32 & $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
-parareal & zia.cerit-sc.cz$^{\mathrm{d}}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
-serial & zia.cerit-sc.cz$^{\mathrm{d}}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
+%\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
+%\midrule[.025em]
+parareal & $10^4$ days & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
+parareal & $10^4$ days & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
+parareal & $10^4$ days & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
+parareal & $10^4$ days & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
+serial & $10^4$ days & 1 & N/A & $32\times10^5$ & 0.003125 day & N/A & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
 \bottomrule
-\multicolumn{11}{l}{$^{\mathrm{a}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
-\multicolumn{11}{l}{$^{\mathrm{b}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
-\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum relative error across the trajectory compared to the serial computation} \\
-\multicolumn{11}{l}{$^{\mathrm{d}}$ The programs were run on the machines of the zia.cerit-sc.cz cluster } \\
-\multicolumn{11}{l}{owned by CERIT-SC/MU, a part of the MetaCentrum organization,} \\
-\multicolumn{11}{l}{each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
+\multicolumn{11}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
+\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
+\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
+\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
 \end{tabular}
@@
 \end{table*}
-\end{comment}
 
 \section{Summary and discussion}
+
+\subsection{Speed-up}
+We measured the run times of a fixed simulation with different numbers of CPUs and compared them with the theoretical expectations.
+As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).
+The implementation is consistently slower than the estimated run times, however, there are many factors that weren't considered in the speed-up estimation, for example we only took the function evaluations in the solvers into account.
+There are also other factors like synchronization costs of the OpenMP threads.
+
+\subsection{Molecular dynamics}
+We evaluated parareal only in the example of our solar system.
+Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate and require very small step sizes.
+
+This is not present in molecular dynamics settings and thus we could expect more stable behavior. This would be an interesting area to continue with this work.
+
+We could further improve our simulations by further specializing them for solar system dynamics as was done in \cite{parallelsolar} (use a different coarse model simulating only the interactions of Sun and other planets).
+However, the direction of this paper is more about general particle simulations, so only the parareal integration was used.
+
+Another particularly interesting method that could be used for molecular dynamics is also the symplectic parareal method described by \cite{symplecticparareal}, mentioned in an earlier section of this paper.
 
@@
 
-I would like to thank my advisor, Samuel J. Newcome, M. Sc., for suggesting directions and providing valueable feedback.
+I would like to thank my advisor, Samuel J. Newcome, M. Sc., for suggesting directions, providing valueable feedback and organising this amazing seminar.
 
@@
 \begin{thebibliography}{00}
-\bibitem{parareal}Bal, G.
-\& Maday, Y.
-A “Parareal” Time Discretization for Non-Linear PDE's with Application to the Pricing of an American Put.
-{\em Recent Developments In Domain Decomposition Methods}.
-pp.
-189-202 (2002)
-6-578 (2007), [Online]. Available: https://doi.org/10.1007/978-3-642-56118-4\_12
+\bibitem{parareal}Lions, J., Maday, Y.
+\& Turinici, G.
+Résolution d'EDP par un schéma en temps «pararéel ».
+{\em Comptes Rendus De L'Académie Des Sciences - Series I - Mathematics}.
+\textbf{332}, 661-668 (2001), [Online].
+Available: \url{https://www.sciencedirect.com/science/article/pii/S0764444200017936}
 \bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
@@
 \textbf{114} pp.
-409 (1997,7), http://dx.doi.org/10.1086/118485
+409 (1997,7), \url{http://dx.doi.org/10.1086/118485}
 \bibitem{speedup}Pentland, K., Tamborrino, M., Sullivan, T., Buchanan, J.
@@
 33, 23 (2022,12), [Online].
-Available: https://doi.org/10.1007/s11222-022-10195-y
+Available: \url{https://doi.org/10.1007/s11222-022-10195-y}
 \bibitem{farfuture}Gander, M.
@@
 21 (2019), [Online].
-Available: http://publications.mfo.de/handle/mfo/3693
+Available: \url{http://publications.mfo.de/handle/mfo/3693}
 \bibitem{parareal2}Gander, M.
@@
 29, 556-578 (2007), [Online].
-Available: https://doi.org/10.1137/05064607X
+Available: \url{https://doi.org/10.1137/05064607X}
 \end{thebibliography}

### 2024-12-05 paper/fig_3.py

@@ -8,7 +8,7 @@ import ref.planets as planets
 
 
 def fig_3():
-	data = np.loadtxt("results/m.32.out.txt")
+	data = np.loadtxt("results/out.32.32.zia.10^4.txt")
 	data, energy = data[:, :-2], data[:, -2:]
 
 	data = data.reshape((data.shape[0], data.shape[1]//3, 3))

### 2024-12-05 paper/fig_4.py

@@ -11,15 +11,19 @@ import ref.planets as planets
 
 
 def fig_4():
-	data = np.loadtxt("results/m.32.out.txt")
+	data = np.loadtxt("results/out.32.32.zia.10^4.txt")
 	data, energy = data[:, :-2], data[:, -2:]
 
+	log_interval = 100
+	dt = 1e4/32e5
+	year = 365.242199
+
 	ax = plt.axes()
-	ax.plot(np.arange(len(energy))*100, energy[:, 0], label="kinetic energy")
-	ax.plot(np.arange(len(energy))*100, energy[:, 1], label="potential energy")
-	ax.plot(np.arange(len(energy))*100, energy.sum(axis=1), label="total energy")
+	ax.plot(np.arange(len(energy))*log_interval * dt / year, energy[:, 0], label="kinetic energy")
+	ax.plot(np.arange(len(energy))*log_interval * dt / year, energy[:, 1], label="potential energy")
+	ax.plot(np.arange(len(energy))*log_interval * dt / year, energy.sum(axis=1), label="total energy")
 	ax.legend()
-	plt.xlabel("Time [day]")
+	plt.xlabel("Time [year]")
 	plt.ylabel(r"Energy $\left[\textrm{M}_{\textrm{Earth}}\frac{\textrm{AU}^2}{\textrm{day}^2}\right]$")
 
 	plt.tight_layout()

### 2024-12-05 paper/fig_5.py

@@ -1,0 +1,26 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+
+import numpy as np
+
+
+def fig_5():
+	ncpus = [1, 4, 16, 32, 64]
+	times = [5*60+1, 2*60+54, 1*60+1, 52, 35]
+
+	xs = np.arange(1, 80)
+	ys = (1/xs + 1/20) * times[0]
+	plt.plot(xs, ys, "r")
+
+	plt.scatter(ncpus[:1], times[:1], color="C3")
+	plt.scatter(ncpus[1:], times[1:], color="C0")
+
+	plt.ylabel("Runtime [s]")
+	plt.xlabel("n CPUs")
+
+	plt.tight_layout()
+	plt.savefig("fig_5.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_5()

### 2024-12-05 parareal/jobs/long.zia.sh
Long sim: attempt 2

@@ -1,15 +1,15 @@
 #!/bin/bash
 #PBS -N parareal_long_test
 #PBS -l select=1:ncpus=32:ompthreads=32:mem=1gb:cluster=zia
-#PBS -l walltime=00:15:00
+#PBS -l walltime=01:45:00
 
 DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 
 # config
 NSEG=32
 MAX_ITERS=12
-COARSE_STEPS=100000
-FINE_STEPS=1000000
+COARSE_STEPS=1000000
+FINE_STEPS=10000000
 TIME=365250
 EPS=1e-3
 LOG_PERIOD=1000

### 2024-12-05 parareal/vis.py
Removed vis.py old copy

@@ -1,41 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-import sys
-
-
-if len(sys.argv) <= 1:
-	print("not enough arguments", file=sys.stderr)
-	sys.exit(1)
-
-data = np.loadtxt(sys.argv[1])
-data, energy = data[:, :-2], data[:, -2:]
-
-data = data.reshape((data.shape[0], data.shape[1]//3, 3))
-
-lim = 1.2*max(data[0, :, :].min(), data[0, :, :].max())
-
-ax = plt.axes(projection="3d")
-ax.set_xlim3d(-lim, lim)
-ax.set_ylim3d(-lim, lim)
-ax.set_zlim3d(-lim, lim)
-
-for i in range(data.shape[1]):
-	ax.plot3D(data[:, i, 0], data[:, i, 1], data[:, i, 2], label=f"particle {i}")
-ax.legend()
-plt.show()
-
-ax = plt.axes()
-ax.plot(np.arange(len(energy)), energy[:, 0], label="kinetic energy")
-ax.plot(np.arange(len(energy)), energy[:, 1], label="potential energy")
-ax.plot(np.arange(len(energy)), energy.sum(axis=1), label="total energy")
-ax.legend()
-plt.show()
-
-ax = plt.axes()
-ax.set_xlim([-lim, lim])
-ax.set_ylim([-lim, lim])
-particles_plot = ax.scatter(data[0, :, 0], data[0, :, 1])
-
-for i in range(len(data)):
-	particles_plot.set_offsets(np.c_[data[i, :, 0], data[i, :, 1]])
-	plt.pause(0.01)

### 2024-12-05 paper/main.tex
Paper: edits

@@
 The purpose of this paper is to examine the applications of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
-We then test and measure the real speed-up and compare it with the expectation derived from theory.
+We then test and measure the real speed-up and compare it with the expectations derived from theory.
 \end{abstract}
@@
 This could be very useful for running smaller simulations through a very long timespan.
-This is of interest for example in solar system dynamics.
-If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time.
-Thus, parallel-in-time algorithms can be employed.\cite{parallelsolar}
+This is of interest for example in solar system dynamics.\footnote{Also in other fields such as molecular dynamics \cite{moldyn} but solar system simulations are a relatively good example.} If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time.
+Thus, parallel-in-time algorithms can be employed.
+\cite{parallelsolar}
 
-In particle simulations, we are often numerically solving an ODE in the form:
+In particle simulations, we are often under the hood numerically solving an ODE in the form:
 $$
@@
 
-The parareal algorithm was introduced by \cite{parareal}.\footnote{The original paper in French, however it includes a description in English.} It achieves parallel-in-time integration using two types of integrators.
+The parareal algorithm was introduced by Lions, Maday and Turinici in 2001.
+\cite{parareal}\footnote{The original paper is in French but it does include a basic description in English.} It achieves parallel-in-time integration using two types of integrators.
 A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to fill in the gaps in this estimated trajectory.
@@
 \subsection*{Stopping criteria}
-We assume the algorithm has converged enough when $\forall n: |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist.\cite{speedup}
+We assume the algorithm has converged enough when $\forall n\colon |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist. \cite{speedup}
 
@@
 
-Thus, to reach a high speed-up, we want the ratio $\frac{T_\mathcal{C}}{T_\mathcal{F}}$ to be small while also keeping the number of required iterations $k$ small.\cite{speedup}
+Thus, to reach a high speed-up, we want the ratio $\frac{T_\mathcal{C}}{T_\mathcal{F}}$ to be small while also keeping the number of required iterations $k$ small. \cite{speedup}
 
-The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up.\cite{speedup}
+The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up.
+\cite{speedup}
 
@@
 $$
-where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from the Newton's law of universal gravitation.
+where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from Newton's law of universal gravitation.
 
 It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver.
-This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.\cite{parallelsolar}\cite{farfuture}
+This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.
+\cite{parallelsolar} \cite{farfuture}
 
@@
 \subsection{Implementation}
-An implementation in C++ using OpenMP for the parallelization was written.
-In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI.
-However, due to the higher complexity and difficulty of such an implementation, an easier approach using OpenMP was used in this project for the present time instead.
+An implementation in C++ using OpenMP for the parallelization was written.\footnote{The code is available at \url{https://github.com/lesves/mol-dyn}.} In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI.
+However, due to the higher complexity and difficulty of such  implementation, an simpler approach with OpenMP was used in this project instead.
 
@@
 
+The experimental runs were conducted on the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization.
+Each computer in the cluster is equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.
+
 A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration).
 
-The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.
+The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figure \ref{sim} and \ref{energy}.
 
@@
 \centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
-\caption{Conservation of energy in the simulation. Same simulation as in Fig. \ref{sim}.}
+\caption{Conservation of energy in the simulation.
+Same simulation as in Figure \ref{sim}.
+With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of timesteps.}
 \label{energy}
@@
 \centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
-\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPUs (blue points).}
+\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPUs (blue points).
+For more information, see Table \ref{runtimes}.}
 \label{speeds}
@@
 \caption{Comparison of running times taken by different configurations of the simulator. 
-$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively. 
-To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
+$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively and $\Delta t_{\text{Final}}$ is the resulting step size.
 The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
@@
 \toprule
-Program & Time & n CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_{\text{Final}}$ & Iters $^{\mathrm{b}}$ & Abs.
+Program & Time & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_{\text{Final}}$ & Iters $^{\mathrm{b}}$ & Abs.
 err.
@@
 As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).
+
 The implementation is consistently slower than the estimated run times, however, there are many factors that weren't considered in the speed-up estimation, for example we only took the function evaluations in the solvers into account.
-There are also other factors like synchronization costs of the OpenMP threads.
+There are also other factors such as the synchronization costs of the OpenMP threads.
 
-\subsection{Molecular dynamics}
+In conclusion, we were able to achieve noticable speed-ups consistent with the theoretical foundation by using the parareal algorithm.
+
+\subsection{Other particle simulations: Molecular dynamics}
 We evaluated parareal only in the example of our solar system.
@@
 
-This is not present in molecular dynamics settings and thus we could expect more stable behavior. This would be an interesting area to continue with this work.
+This is usually not present in molecular dynamics settings and thus we could expect a more stable behavior. This could be an interesting area to continue with this work.
 
-We could further improve our simulations by further specializing them for solar system dynamics as was done in \cite{parallelsolar} (use a different coarse model simulating only the interactions of Sun and other planets).
-However, the direction of this paper is more about general particle simulations, so only the parareal integration was used.
+On the other hand, we could further improve our solar system simulations by further specializing them as was done in \cite{parallelsolar} (a different coarse model simulating only the interactions of Sun and other planets was used instead of a course integrator).
+However, our intent is to study general particle simulations, so this would be beyond our scope.
 
@@
 
-\section*{Acknowledgment}
-\addcontentsline{toc}{section}{Acknowledgment}
+\section*{Acknowledgements}
+\addcontentsline{toc}{section}{Acknowledgements}
 
@@
 Available: \url{https://www.sciencedirect.com/science/article/pii/S0764444200017936}
-\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
 \bibitem{parallelsolar}Saha, P., Stadel, J.
@@
 Available: \url{https://doi.org/10.1007/s11222-022-10195-y}
+\bibitem{parareal2}Gander, M.
+\& Vandewalle, S.
+``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing.
+29, 556-578 (2007), [Online].
+Available: \url{https://doi.org/10.1137/05064607X}
 \bibitem{farfuture}Gander, M.
@@
 Available: \url{http://publications.mfo.de/handle/mfo/3693}
-\bibitem{parareal2}Gander, M.
-\& Vandewalle, S.
-``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing.
-29, 556-578 (2007), [Online].
-Available: \url{https://doi.org/10.1137/05064607X}
+\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
+\bibitem{moldyn}Baffico, L., Bernard, S., Maday, Y., Turinici, G.
+\& Zérah, G.
+Parallel-in-time molecular-dynamics simulations.
+{\em Physical Review E}.
+\textbf{66} (2002,11), http://dx.doi.org/10.1103/physreve.66.057701
+
 \end{thebibliography}

### 2025-01-28 parareal/jobs/job.serial.prec2.zia.10^4.sh
Lower precision serial test

@@ -1,0 +1,21 @@
+#!/bin/bash
+#PBS -N parareal_serial_test
+#PBS -l select=1:ncpus=1:ompthreads=1:mem=1gb:cluster=zia
+#PBS -l walltime=00:06:00
+
+DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
+
+# config
+NSEG=32
+MAX_ITERS=16
+COARSE_STEPS=10000 # COARSE_STEPS does nothing
+FINE_STEPS=50000 # We intentionally lower the number of steps to decrease precision
+TIME=10000
+EPS=1e-3
+LOG_PERIOD=100
+
+# run
+OUT="out.serial.prec2.zia.10^4.txt"
+FLAGS="--serial"
+
+source $DATADIR/jobs/base_job.sh

### 2025-01-28 parareal/jobs/job.serial.prec3.zia.10^4.sh
Lower precision serial tests

@@ -9,13 +9,13 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 NSEG=32
 MAX_ITERS=16
 COARSE_STEPS=10000 # COARSE_STEPS does nothing
-FINE_STEPS=50000 # We intentionally lower the number of steps to decrease precision
+FINE_STEPS=90000 # We intentionally lower the number of steps to decrease precision
 TIME=10000
 EPS=1e-3
 LOG_PERIOD=100
 
 # run
-OUT="out.serial.prec2.zia.10^4.txt"
+OUT="out.serial.prec4.zia.10^4.txt"
 FLAGS="--serial"
 
 source $DATADIR/jobs/base_job.sh

### 2025-01-28 parareal/jobs/job.serial.prec4.zia.10^4.sh

@@ -9,13 +9,13 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 NSEG=32
 MAX_ITERS=16
 COARSE_STEPS=10000 # COARSE_STEPS does nothing
-FINE_STEPS=50000 # We intentionally lower the number of steps to decrease precision
+FINE_STEPS=75000 # We intentionally lower the number of steps to decrease precision
 TIME=10000
 EPS=1e-3
 LOG_PERIOD=100
 
 # run
-OUT="out.serial.prec2.zia.10^4.txt"
+OUT="out.serial.prec3.zia.10^4.txt"
 FLAGS="--serial"
 
 source $DATADIR/jobs/base_job.sh

### 2025-01-28 parareal/jobs/job.serial.prec3.zia.10^4.sh
Fixed swapped file names

@@ -9,13 +9,13 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 NSEG=32
 MAX_ITERS=16
 COARSE_STEPS=10000 # COARSE_STEPS does nothing
-FINE_STEPS=90000 # We intentionally lower the number of steps to decrease precision
+FINE_STEPS=75000 # We intentionally lower the number of steps to decrease precision
 TIME=10000
 EPS=1e-3
 LOG_PERIOD=100
 
 # run
-OUT="out.serial.prec4.zia.10^4.txt"
+OUT="out.serial.prec3.zia.10^4.txt"
 FLAGS="--serial"
 
 source $DATADIR/jobs/base_job.sh

### 2025-01-28 parareal/jobs/job.serial.prec4.zia.10^4.sh

@@ -9,13 +9,13 @@ DATADIR=/storage/praha1/home/skywalker/mol-dyn/parareal
 NSEG=32
 MAX_ITERS=16
 COARSE_STEPS=10000 # COARSE_STEPS does nothing
-FINE_STEPS=75000 # We intentionally lower the number of steps to decrease precision
+FINE_STEPS=90000 # We intentionally lower the number of steps to decrease precision
 TIME=10000
 EPS=1e-3
 LOG_PERIOD=100
 
 # run
-OUT="out.serial.prec3.zia.10^4.txt"
+OUT="out.serial.prec4.zia.10^4.txt"
 FLAGS="--serial"
 
 source $DATADIR/jobs/base_job.sh

### 2025-01-28 paper/main.tex
Added tests regarding number of iterations, run time and the number of coarse steps

@@
 \usepackage{xurl}
+\usepackage{pbalance}
 
@@
 
-\title{Parallel-in-time methods for Particle Simulations}
+\title{Parallel-in-time Methods for Particle Simulations}
 
 \author{\IEEEauthorblockN{Lukáš Veškrna}
-\IEEEauthorblockA{\textit{Technical University Munich} \\
+\IEEEauthorblockA{\textit{Technical University Munich}\\ \textit{\& Charles University} \\
 Munich, Germany \\
@@
 \begin{abstract}
-The purpose of this paper is to examine the applications of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
-We then test and measure the real speed-up and compare it with the expectations derived from theory.
+The purpose of this article is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
+We present the results of experiments measuring the run times, which we then compare with expectations derived from theory.
 \end{abstract}
@@
 
-There do, however, also exist algorithms that allow us to parallelize simulations in the time domain.
-This could be very useful for running smaller simulations through a very long timespan.
-This is of interest for example in solar system dynamics.\footnote{Also in other fields such as molecular dynamics \cite{moldyn} but solar system simulations are a relatively good example.} If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time.
-Thus, parallel-in-time algorithms can be employed.
-\cite{parallelsolar}
+There do, however, also exist algorithms that allow us to parallelize simulations in the time domain. This could be very useful for running smaller simulations through a very long timespan.
+
+In this article, we will focus on the fitting example of solar system dynamics, although many applications could be found also in other fields, such as molecular dynamics\cite{moldyn}.
+
+If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time. Thus, parallel-in-time algorithms can be employed. \cite{parallelsolar}
 
@@
 Supposing we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta T = \frac{T-t_0}{N}$.
-We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta T$ (so $t_0 = t_0$ and $t_N = T$).
+We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta T$ (so $t_0 = t_0$ and $t_N = T$).\footnote{The solvers can also use their own smaller subdivisions (e.g.
+different step sizes) but the parareal algorithm does not interact with those subdivisions.}
 
@@
 
-In the 0th iteration, we use the coarse solver to compute $u_1\dots u_n$ sequentially. For every $0 \leq n < N$ we do
+In the 0th iteration, we use the coarse solver to compute $u_1\dots u_N$ sequentially. For every $0 \leq n < N$ we do
 $$
@@
 An implementation in C++ using OpenMP for the parallelization was written.\footnote{The code is available at \url{https://github.com/lesves/mol-dyn}.} In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI.
-However, due to the higher complexity and difficulty of such  implementation, an simpler approach with OpenMP was used in this project instead.
+However, due to the higher complexity and difficulty of such  implementation, a simpler approach with OpenMP was used in this project instead.
 
-\subsection{Experiments (for run time comparison)}
-Multiple experiments with the same initial conditions and same step size were ran on a cluster in order to get a more reproducible and comparable environment.
+\subsection{Experiments: The environment}
+All experiments were ran on a cluster in order to get a more reproducible and comparable environment.
 
@@
 
-A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration).
+\subsection{Experiment I: Measured run time and expected run time comparison}
+
+Multiple experiments with the same initial conditions and same step size were conducted.
+A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration\footnote{This choice is explained in the next experiment (Experiment II), where the observations suggest that convergence within one iteration can be the situation with optimal run time.}).
+For comparison, also a serial version (not using the parareal algorithm) was ran with the same initial conditions and step size.
 
@@
 \centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
-\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see Table \ref{runtimes}.}
+\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPU cores. For more information, see Table \ref{runtimes}.}
 \label{sim}
@@
 
-\subsection{Run time comparison}
-Let's derive the expected speed-up and run time of our program and compare it with the measured run time.
+Let's derive the expected speed-ups and run times of our program and compare it with the measured run times.
 To simplify the calculation, we will only estimate it using the number of function evaluations.
@@
 \centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
-\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPUs (blue points).
+\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPU cores (blue points).
 For more information, see Table \ref{runtimes}.}
@@
 
+\subsection{Experiment II: Run times, numbers of iterations and coarse solver steps}
+
+The second experiment shows how do run times and the numbers of iterations that the program takes to converge relate to the number of steps of the coarse solver.
+This observation can help with deciding what is the optimal number of steps to use for the coarse solver.
+We are interested primarily in the run time, however the number of iterations turns out to be a very important factor in the run time.
+
+A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were ran and the results are shown in figure \ref{timetest}.
+The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.
+
+A key feature observed in figure \ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes.
+After every initial downward spike, there is always an upward trend.
+
+Presumably, this shows that the run time depends primarily on the number of iterations and only secondarily on the number of steps of the coarse solver.
+The upward trends correspond to improving the coarse solver while not improving the number of iterations, which does not improve the computation time.
+
+From this, we can conjecture, that the optimal choice of the number of steps for the coarse solver is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).
+
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=\linewidth]{fig_6.eps}}
+\caption{Run time and number of iterations depending on the ratio of the numbers of steps of the coarse and fine solvers.
+It corresponds to a simulation of two years with final step size $\frac{2\times365\text{ day}}{800\times 16} \approx 0.057\text{ day}$ run on 16 cores (with $N_\mathcal{F} = 800$, $\mathcal{F}$ is once again RK4 and $\mathcal{C}$ is Velocity Verlet).
+It was averaged through 10 runs to reduce noise interfering with the small running times.
+The $\varepsilon$ for convergence was chosen as $\varepsilon = 10^{-3}$.}
+\label{timetest}
+\end{figure}
+
 \begin{table*}[htbp]
-\caption{Comparison of running times taken by different configurations of the simulator. 
-$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively and $\Delta t_{\text{Final}}$ is the resulting step size.
+\caption{Comparison of running times taken by different configurations of the simulator for a planetary simulation of $10^4$ days.
+$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively.
+The fine solver step numbers have been chosen specifically to preserve the step size of 0.003125 day.
+This step size was selected, so all the parareal runs converge within one iteration (with $\varepsilon = 10^{-3}\text{ AU}$).
+This number of iterations was chosen for easier comparison between the runs and it is a low number to achieve greater speed-up.
 The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
 \begin{center}
-\begin{tabular}{crrrrrrrrrcc}
+\begin{tabular}{crrrrrrrrcc}
 \toprule
-Program & Time & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_{\text{Final}}$ & Iters $^{\mathrm{b}}$ & Abs.
-err.
-$^{\mathrm{c}}$ & Rel.
-err.
-$^{\mathrm{d}}$ & CPU time & Real time \\
+Program & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & Abs. err. $^{\mathrm{b}}$ & Rel. err. $^{\mathrm{c}}$ & CPU time & Real time \\
 \midrule
@@
 %\midrule[.025em]
-parareal & $10^4$ days & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
-parareal & $10^4$ days & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
-parareal & $10^4$ days & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
-parareal & $10^4$ days & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
-serial & $10^4$ days & 1 & N/A & $32\times10^5$ & 0.003125 day & N/A & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
+parareal & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
+parareal & 32 & $1\times10^4$ &  $1\times10^5$ & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
+parareal & 16 & $2\times10^4$ &  $2\times10^5$ & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
+parareal & 4 & $8\times10^4$ &  $8\times10^5$ & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
+serial & 1 & N/A & $32\times10^5$ & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
 \bottomrule
-\multicolumn{11}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
-\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
-\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
-\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
+\multicolumn{8}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
+\multicolumn{8}{l}{$^{\mathrm{b}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
+\multicolumn{8}{l}{$^{\mathrm{c}}$ Maximum relative error across the trajectory compared to the serial computation}
 \end{tabular}
@@
 \subsection{Speed-up}
-We measured the run times of a fixed simulation with different numbers of CPUs and compared them with the theoretical expectations.
+We measured the run times of a fixed simulation with different numbers of CPU cores and compared them with the theoretical expectations.
 As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).
@@
 In conclusion, we were able to achieve noticable speed-ups consistent with the theoretical foundation by using the parareal algorithm.
+
+\subsection{Spatial vs temporal parallelization}
+What configurations of the simulation size and the timespan benefit from parallel-in-time integration?
+How does spatial and temporal parallelization combine together?
+This is beyond scope of this article, however {\color{red} TODO}
 
@@
 
-I would like to thank my advisor, Samuel J. Newcome, M. Sc., for suggesting directions, providing valueable feedback and organising this amazing seminar.
+I would like to thank my advisor, Samuel J.
+Newcome, M.
+Sc.
+and all other participants of the Seminar in Methods for Molecular Dynamics, for suggesting directions, providing valueable feedback and organising this amazing seminar.
 

### 2025-01-28 paper/fig_6.py

@@ -1,0 +1,24 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+
+import numpy as np
+
+
+def fig_6():
+	ratios, iters, times = np.loadtxt("results/timetest-result.txt").reshape((3, -1))
+
+	fig, ax = plt.subplots(2,)
+	#times_avg = np.convolve(times, np.ones(WINDOW_SIZE), "valid") / WINDOW_SIZE
+
+	ax[0].plot(ratios, iters)
+	ax[0].set_xlabel(r"$\frac{N_\mathcal{C}}{N_\mathcal{F}}$")
+	ax[0].set_ylabel(r"No. of iterations")
+	ax[1].plot(ratios, times)
+	#ax[1].plot(ratios[WINDOW_SIZE-1:], times_avg)
+	ax[1].set_xlabel(r"$\frac{N_\mathcal{C}}{N_\mathcal{F}}$")
+	ax[1].set_ylabel(r"Run time [s]")
+	plt.savefig("fig_6.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_6()

### 2025-01-28 scripts/timetest.py

@@ -22,7 +22,7 @@ EPS = 1e-3
 
 COARSE_RANGE = range(0, FINE+1, 8)
 
-N_TRIALS = 10
+N_TRIALS = 1
 
 
 def run_experiment():

### 2025-02-01 paper/main.tex
New version of the paper; Figure 7

@@
 \usepackage{xurl}
-\usepackage{pbalance}
+
+% Character support
+\usepackage[english]{babel}
+
+% Turn off small-caps in table caption
+% I don't like it, I want to have a long caption
+% and this doesn't look good with it
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd{\@makecaption}
+  {\scshape}
+  {}
+  {}
+  {}
+\makeatother
+
+%\renewcommand{\sfdefault}{cmss}
+%\renewcommand{\rmdefault}{cmr}
+%\renewcommand{\ttdefault}{cmt}
 
@@
 \begin{abstract}
-The purpose of this article is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
-We present the results of experiments measuring the run times, which we then compare with expectations derived from theory.
+The purpose of this paper is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
+We present the results of experiments measuring the run times of the simulation, which we then compare with expectations derived from theory.
 \end{abstract}
@@
 
-There do, however, also exist algorithms that allow us to parallelize simulations in the time domain. This could be very useful for running smaller simulations through a very long timespan.
+There do, however, also exist algorithms that allow us to parallelize simulations in the time domain.
+This could be very useful for running smaller simulations through a very long timespan or with a very small step size.
 
-In this article, we will focus on the fitting example of solar system dynamics, although many applications could be found also in other fields, such as molecular dynamics\cite{moldyn}.
+In this article, we will focus on the fitting example of solar system dynamics, although many applications could be found also in other fields, such as molecular dynamics. \cite{moldyn}
 
@@
 $$
-where $f$ is the function describing our problem, $u(t)$ is the state of the system at time $t$ and we are given an initial condition $u(t_0) = u_0$.
+where $f$ is the function describing our problem, $u(t)$ is the state of the system at time $t$ and we are given an initial condition $u(t_0) = u_0$ (this type of problem is often called an initial value problem).
 
-The integration part is the part which usually is dependent on some previous time step and thus can't be easily parallelized.
+We typically solve this problem using a numerical method, that does a step in time of size $\Delta t$ and estimates $u(t_0+\Delta t)$.
+We then use this result to advance the computation further in time by taking more such steps.
+
+Thus, the integration is the part of the computation which usually is dependent on some previous time step and thus can't be easily parallelized.
 In order to make it parallelizable, multiple techniques were devised, one of them being the parareal algorithm.
 
-\section{The parareal algorithm}
+\section{The Parareal Algorithm}
 
@@
 \cite{parareal}\footnote{The original paper is in French but it does include a basic description in English.} It achieves parallel-in-time integration using two types of integrators.
-A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to fill in the gaps in this estimated trajectory.
+A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution\footnote{In the literature, the coarse solver is often denoted $\mathcal{G}$.
+However, I find it more intuitive to use $\mathcal{C}$ as in the word \textbf{c}oarse.}, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to further refine this estimated trajectory.
 
-Supposing we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta T = \frac{T-t_0}{N}$.
-We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta T$ (so $t_0 = t_0$ and $t_N = T$).\footnote{The solvers can also use their own smaller subdivisions (e.g.
+Supposing we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta t = \frac{T-t_0}{N}$.
+We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta t$ (so $t_0 = t_0$ and $t_N = T$).\footnote{The solvers can also use their own smaller subdivisions (e.g.
 different step sizes) but the parareal algorithm does not interact with those subdivisions.}
 
-The algorithm works in iterations. The 0th iteration is used to initialize the algorithm, while the subsequent iterations improve the result of the previous iteration.
+The algorithm works in iterations.
+The 0th iteration is used to initialize the algorithm, while the subsequent iterations improve the result of the previous iteration.
+We use the notation $u^i_n$ to represent $u_n$ for iteration $i$.
 
@@
 $$
-u^0_{n+1} = \mathcal{C}(t_n, t_n+\Delta T, u^0_n)
+u^0_{n+1} = \mathcal{C}(t_n, t_n+\Delta t, u^0_n)
 $$
@@
 
-In every subsequent iteration $i+1$, we first calculate the results of the fine but expensive solver, $\mathcal{F}(t_n, t_n+\Delta T, u^{i}_n)$, for every $0 \leq n < N$ in parallel.
+In every subsequent iteration $i+1$, we first calculate the results of the fine but expensive solver, $\mathcal{F}(t_n, t_n+\Delta t, u^{i}_n)$, for every $0 \leq n < N$ in parallel.
 
@@
 \begin{multline}
-u^{i+1}_{n+1} = \mathcal{C}(t_n, t_n+\Delta T, u^{i+1}_n)\\ + \mathcal{F}(t_n, t_n+\Delta T, u^i_n)\\ - \mathcal{C}(t_n, t_n+\Delta T, u^{i}_n)
+u^{i+1}_{n+1} = \mathcal{C}(t_n, t_n+\Delta t, u^{i+1}_n)\\ + \mathcal{F}(t_n, t_n+\Delta t, u^i_n)\\ - \mathcal{C}(t_n, t_n+\Delta t, u^{i}_n)
 \end{multline}
@@
 This is repeated until satisfactory convergence.
-It has been proven for example by \cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
+It has been proven for example by \cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta t, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
 
@@
 \centerline{\includegraphics[width=\linewidth]{fig_1.eps}}
-\caption{Parareal convergence for the example of a harmonic oscillator with $k = 1\frac{\text{N}}{\text{m}}$. The coarse solver is Euler and the fine solver is RK4.}
+\caption{Parareal convergence for the example of a harmonic oscillator with $k = 1\frac{\text{N}}{\text{m}}$. The coarse solver is explicit Euler and the fine solver is RK4.}
 \label{oscillator}
@@
 \subsection*{Stopping criteria}
-We assume the algorithm has converged enough when $\forall n\colon |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist. \cite{speedup}
+We assume the algorithm has converged enough when $\forall n\colon \|u^{i+1}_n - u^i_n\| < \varepsilon$. Other criteria also exist. \cite{speedup}
 
@@
 
-A derivation of the theoretical speedup is presented in \cite{speedup} (Section 2.3), which we use here.
+A derivation of the theoretical speedup is presented in Pentland et al.
+\cite{speedup} (Section 2.3), which we use here.
 It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially (our implementation doesn't include this optimization).
 
-To find the theoretical speed-up of the algorithm, let's compute the time required by the parareal algorithm, $T_\text{parareal}$, and the time $T_\text{serial}$ taken to compute the solution by running the fine integrator serially.
+To find the theoretical speed-up of the algorithm, we compute the time required by the parareal algorithm, $T_\text{parareal}$, and the time $T_\text{serial}$ taken to compute the solution by running the fine integrator serially.
 
-Assume running $\mathcal{F}$ over any segment $[t_n, t_{n+1}]$ takes $T_\mathcal{F}$ and running $\mathcal{C}$ over any segment takes $T_\mathcal{C}$. Then computing $\mathcal{F}$ serially takes
+If we assume running $\mathcal{F}$ over any segment $[t_n, t_{n+1}]$ takes $T_\mathcal{F}$ and running $\mathcal{C}$ over any segment takes $T_\mathcal{C}$, then computing $\mathcal{F}$ serially takes
 $$
@@
 $$
+where $k$ is the number of iterations required for convergence.
 
@@
 
-The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up.
-\cite{speedup}
+The algorithm always converges in at most $N$ iterations, because the exact initial condition propagates through all the $N$ segments, however, that is not useful (and takes even more time), because the algorithm has to converge at a significantly smaller number of iterations to gain any real speed-up.
+\cite{speedup} (2.3)
 
-\section{Application to particle simulations}
+\section{Application to Particle Simulations}
 The purpose of this paper is to examine the potential in the use of the parareal method for long running simulations such as simulations of the solar system.
@@
 We use a simple $n$-body particle simulation of the solar system with time complexity $\mathcal{O}(n^2)$, which we then integrate using the parareal method.
-This is inefficient for a large number of particles, however, our focus is on the parallel-in-time integration, therefore we use a simpler simulation.
+This is inefficient for a large number of particles, however, our focus is on the parallel-in-time integration, not the optimizations required for a large number of particles, therefore we use a simpler simulation.
 
@@
 $$
-\diff{\mathbf{u}(t)}{t} = f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
+\diff{u(t)}{t} = f(t, u(t)) = f(t, (x(t), v(t))) = (v(t), a(t))
 $$
@@
 It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver.
-This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.
+This was done prior to the invention of parareal by Saha et al.
+\cite{parallelsolar}, also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.
 \cite{parallelsolar} \cite{farfuture}
@@
     \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties with a small cost and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision.
+\cite{integration}
     \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic.
@@
 
+For completeness, the two selected integration methods are briefly described in the next paragraphs.
+\subsubsection{Velocity Verlet}
+
+\begin{align*}
+x(t+\Delta t) &= x(t) + v(t)\Delta t + \frac{1}{2}a(t)\Delta t^2 \\
+v(t+\Delta t) &= v(t) + \frac{a(t) + a(t+\Delta t)}{2}\Delta t
+\end{align*}
+Here $x(t)$, $v(t)$ and $a(t)$ represent approximations of the corresponding values (position, velocity, acceleration) at timestep $t$, $\Delta t$ is the time step size.
+Note that only one evaluation of $a$ per step suffices, because we can keep $a(t)$ from the last step.
+\cite{forverlet}
+
+\subsubsection{The classic fourth-order Runge-Kutta method (RK4)}
+\begin{align*}
+k_1 &= f(t, u(t)) \\
+k_2 &= f(t + \frac{\Delta t}{2}, u(t) + \Delta t\frac{k_1}{2}) \\
+k_3 &= f(t + \frac{\Delta t}{2}, u(t) + \Delta t\frac{k_2}{2}) \\
+k_4 &= f(t + \Delta t, u(t) + \Delta t k_3) \\
+u(t+\Delta t) &= u(t) + \frac{\Delta t}{6}\left(k_1 + 2k_2 + 2k_3 + k_4\right)
+\end{align*}
+where $u(t)$ represents the approximation of the state at time $t$. $\Delta t$ denotes the time step size. The method requires four acceleration computations per step. \cite{forrk4}
+
+\subsection{The initial conditions}
+The initial conditions for the simulation were generated with a python script using the astropy package. \cite{astropy}
+
 \begin{figure}[htbp]
 \centerline{\includegraphics[width=\linewidth]{fig_2.eps}}
-\caption{Path of Mercury over the time of one Earth year in a parareal integrated simulation converging over multiple iterations of the parareal algorithm.
+\caption{A more complex example to illustrate the behavior of the parareal algorithm: Path of Mercury over the time of one Earth year in a parareal integrated simulation converging over multiple iterations of the parareal algorithm.
 Velocity verlet and RK4 are used with the same step size of $0.5\text{ day}$.}
@@
 
-\section{Implementation and experiments}
+\section{Implementation and Experiments}
 
@@
 An implementation in C++ using OpenMP for the parallelization was written.\footnote{The code is available at \url{https://github.com/lesves/mol-dyn}.} In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI.
-However, due to the higher complexity and difficulty of such  implementation, a simpler approach with OpenMP was used in this project instead.
+However, due to the higher complexity of such implementation, a simpler approach with OpenMP was used in this project instead.
 
@@
 Multiple experiments with the same initial conditions and same step size were conducted.
-A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration\footnote{This choice is explained in the next experiment (Experiment II), where the observations suggest that convergence within one iteration can be the situation with optimal run time.}).
+A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration\footnote{This choice is explained in the next experiment (\textit{Experiment II}), where the observations suggest that convergence within one iteration possibly has optimal run time.}).
 For comparison, also a serial version (not using the parareal algorithm) was ran with the same initial conditions and step size.
@@
 \centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
-\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPU cores. For more information, see Table \ref{runtimes}.}
+\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal. For more information, see Table \ref{runtimes}.}
 \label{sim}
@@
 \centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
-\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPU cores (blue points).
+\caption{Comparison of the expected run times (red curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red ``x'' point)) and the real measured times with different numbers of CPU cores (blue points).
 For more information, see Table \ref{runtimes}.}
@@
 \end{figure}
+
+While we loose some accuracy with the parareal method, this is a parameter that can be controlled, for example using the converge parameter ($\varepsilon$) and the coarse solver accuracy.
+
+If we instead approximate the simulation using a similar serial simulation with a lower number of steps, we get a substantially bigger error than the parareal algorithm while not improving the run time significantly.
+This is also shown in Table \ref{runtimes} (as ``serial (approx.)'').
 
@@
 
-A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were ran and the results are shown in figure \ref{timetest}.
+A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were ran and the results are shown in Figure \ref{timetest}.
 The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.
 
-A key feature observed in figure \ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes.
+A key feature observed in Figure \ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes.
 After every initial downward spike, there is always an upward trend.
@@
 
-From this, we can conjecture, that the optimal choice of the number of steps for the coarse solver is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).
+From this, we can conjecture, that (in the case of our simulation) the optimal choice of the number of steps for the coarse solver (in the sense of the smallest run time with a fixed $\varepsilon$) is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).
 
@@
 
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=\linewidth]{fig_7.eps}}
+\caption{This bonus figure shows the numbers of iterations that the parareal algorithm with $N = 16$ requires to converge for a simulation of 365 days with the fine solver step size of $\frac{T-t_0}{NN_\mathcal{F}} = \frac{365\text{ days}}{16\times100}$ for different ratios of $\frac{N_\mathcal{C}}{N_\mathcal{F}}$ and different values of the convergence criterion parameter, $\varepsilon$.}
+\label{arraytest}
+\end{figure}
+
 \begin{table*}[htbp]
 \caption{Comparison of running times taken by different configurations of the simulator for a planetary simulation of $10^4$ days.
-$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively.
+$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively (within one segment).
 The fine solver step numbers have been chosen specifically to preserve the step size of 0.003125 day.
 This step size was selected, so all the parareal runs converge within one iteration (with $\varepsilon = 10^{-3}\text{ AU}$).
-This number of iterations was chosen for easier comparison between the runs and it is a low number to achieve greater speed-up.
+This number of iterations was chosen for easier comparison between the runs and to achieve greater speed-up (see \textit{Experiment II}).
 The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
 \begin{center}
-\begin{tabular}{crrrrrrrrcc}
+\begin{tabular}{crrrrrrrrrcc}
 \toprule
-Program & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & Abs. err. $^{\mathrm{b}}$ & Rel. err. $^{\mathrm{c}}$ & CPU time & Real time \\
+Program & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_\mathcal{F}$ $^{\mathrm{b}}$ & Abs. err. $^{\mathrm{c}}$ & Rel. err. $^{\mathrm{d}}$ & CPU time & Real time \\
 \midrule
@@
 %\midrule[.025em]
-parareal & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
-parareal & 32 & $1\times10^4$ &  $1\times10^5$ & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
-parareal & 16 & $2\times10^4$ &  $2\times10^5$ & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
-parareal & 4 & $8\times10^4$ &  $8\times10^5$ & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
-serial & 1 & N/A & $32\times10^5$ & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
+parareal & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
+parareal & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
+parareal & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
+parareal & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
+serial (base) & 1 & N/A & $32\times10^5$ & 0.003125 day & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
+serial (approx.) $^{\mathrm{e}}$ & 1 & N/A & $0.9\times32\times10^5$ & 0.003472 day & 6.7580 AU & 256.25\% & 00:04:38 & 00:04:38 \\
 \bottomrule
 \multicolumn{8}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
-\multicolumn{8}{l}{$^{\mathrm{b}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
-\multicolumn{8}{l}{$^{\mathrm{c}}$ Maximum relative error across the trajectory compared to the serial computation}
+\multicolumn{8}{l}{$^{\mathrm{b}}$ The resulting step size (fine solver step size). For parareal, it can be calculated as $\frac{T-t_0}{NN_\mathcal{F}}$} \\
+\multicolumn{8}{l}{$^{\mathrm{c}}$ Maximum difference from the base serial computation across the trajectory} \\
+\multicolumn{8}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the base serial computation} \\
+\multicolumn{8}{l}{$^{\mathrm{e}}$ A serial approximation of the base serial simulation using a bigger step size.}
 \end{tabular}
@@
 
-\section{Summary and discussion}
+\section{Summary and Discussion}
 
 \subsection{Speed-up}
-We measured the run times of a fixed simulation with different numbers of CPU cores and compared them with the theoretical expectations.
+The run times of a fixed simulation with different numbers of CPU cores were measured in Experiment I and compared with the theoretical expectations.
 As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).
@@
 
+By comparing parareal with an approximating serial run using a bigger step size (see ''serial (approx.)`` in Table \ref{runtimes}), we have also shown that in this case, the parareal algorithm beats simply approximating the solution using a smaller step size in both accuracy and run time.
+
+The error of the parareal algorithm can be controlled using parameters like the threshold for the convergence criterion ($\varepsilon$) and in some cases the coarse solver step size.
+
+And finally in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation.
+The results are presented in Figure \ref{timetest}.
+The result is a prediction of the optimal coarse solver step size for our simulation, which is the biggest possible such that the algorithm converges in one iteration.
+
 \subsection{Spatial vs temporal parallelization}
-What configurations of the simulation size and the timespan benefit from parallel-in-time integration?
-How does spatial and temporal parallelization combine together?
-This is beyond scope of this article, however {\color{red} TODO}
+
+In some situations it is not that clear as in our case whether is it better to use spatial or temporal parallelization or both.
+This is beyond the scope of this article, however, there has been research in this area.
+
+For example, Croce et al.
+presented a study using a simulation of unsteady Navier-Stokes equations for incompressible flow using parareal that is parallel both in time and space.
+They measured the speed-up with different numbers of cores being asigned to the spatial and temporal parallelization and shown that they can be combined to gain higher speed-up.
+\cite{spacetime}
 
 \subsection{Other particle simulations: Molecular dynamics}
+In molecular dynamics, there are potential applications for parallel-in-time integration, such as material science.
+\cite{moldyn} Or possibly simulating protein folding, which requires extremely small step sizes (as small as femtoseconds) and great amounts of computational power.
+\cite{protein}
+
 We evaluated parareal only in the example of our solar system.
-Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate and require very small step sizes.
+Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate in the sense that if we simulate a single orbit of Neptune, we have to simulate many orbits of Mercury (in a simulation of all the planets), which requires very small step sizes to behave correctly.
 
-This is usually not present in molecular dynamics settings and thus we could expect a more stable behavior. This could be an interesting area to continue with this work.
+We could further improve our solar system simulations by further specializing them as was done in the work by Saha et al.
+\cite{parallelsolar} (a different coarse model simulating only the interactions of Sun and other planets was used instead of a course integrator).
+However, this would be too specific for solar system dynamics and wound not generalize well, so this is beyond our scope.
 
-On the other hand, we could further improve our solar system simulations by further specializing them as was done in \cite{parallelsolar} (a different coarse model simulating only the interactions of Sun and other planets was used instead of a course integrator).
-However, our intent is to study general particle simulations, so this would be beyond our scope.
+These problems are usually not present in molecular dynamics settings and thus we could expect a more stable behavior.
+However, molecular dynamics present their own different challenges.
+This could be an interesting area to continue with this work.
 
-Another particularly interesting method that could be used for molecular dynamics is also the symplectic parareal method described by \cite{symplecticparareal}, mentioned in an earlier section of this paper.
+%Another particularly interesting method that could be used for molecular dynamics is also the symplectic parareal method described by \cite{symplecticparareal}, mentioned in an earlier section of this paper.
 
@@
 
+\IEEEtriggeratref{9}
 \begin{thebibliography}{00}
@@
 \& Turinici, G.
-Résolution d'EDP par un schéma en temps «pararéel ».
+Résolution d'EDP par un schéma en temps «pararéel».
 {\em Comptes Rendus De L'Académie Des Sciences - Series I - Mathematics}.
@@
 \textbf{114} pp.
-409 (1997,7), \url{http://dx.doi.org/10.1086/118485}
+409 (1997,7), [Online].
+Available: \url{http://dx.doi.org/10.1086/118485}
 \bibitem{speedup}Pentland, K., Tamborrino, M., Sullivan, T., Buchanan, J.
@@
 {\em Physical Review E}.
-\textbf{66} (2002,11), http://dx.doi.org/10.1103/physreve.66.057701
-
+\textbf{66} (2002,11), [Online].
+Available: \url{http://dx.doi.org/10.1103/physreve.66.057701}
+\bibitem{integration}Leimkuhler, B., Reich, S.
+\& Skeel, R.
+Integration Methods for Molecular Dynamics.
+{\em Mathematical Approaches To Biomolecular Structure And Dynamics}.
+pp.
+161-185 (1996), [Online].
+Available: \url{https://doi.org/10.1007/978-1-4612-4066-2_10}
+\bibitem{forverlet}Swope, W., Andersen, H., Berens, P.
+\& Wilson, K.
+A computer simulation method for the calculation of equilibrium constants for the formation of physical clusters of molecules: Application to small water clusters.
+{\em The Journal Of Chemical Physics}.
+\textbf{76}, 637-649 (1982,1), [Online].
+Available: \url{https://doi.org/10.1063/1.442716}
+\bibitem{forrk4}Steven C. Chapra, R. Numerical Methods for Engineers. (McGraw-Hill Education,2014)
+\bibitem{spacetime}Croce, R., Ruprecht, D.
+\& Krause, R.
+Parallel-in-Space-and-Time Simulation of the Three-Dimensional, Unsteady Navier-Stokes Equations for Incompressible Flow.
+{\em Modeling, Simulation And Optimization Of Complex Processes - HPSC 2012}.
+pp.
+13-23 (2014)
+\bibitem{protein}Chodera, J., Swope, W., Pitera, J.
+\& Dill, K.
+Long‐Time Protein Folding Dynamics from Short‐Time Molecular Dynamics Simulations.
+{\em Multiscale Modeling \& Simulation}.
+\textbf{5}, 1214-1226 (2006), [Online].
+Available: \url{https://doi.org/10.1137/06065146X}
+\bibitem{astropy}Astropy Collaboration, Price-Whelan, A., Lim, P., Earl, N., Starkman, N., Bradley, L., Shupe, D., Patil, A., Corrales, L., Brasseur, C., Nöthe, M., Donath, A., Tollerud, E., Morris, B., Ginsburg, A., Vaher, E., Weaver, B., Tocknell, J., Jamieson, W., Van Kerkwijk, M., Robitaille, T., Merry, B., Bachetti, M., Günther, H., Aldcroft, T., Alvarado-Montes, J., Archibald, A., Bódi, A., Bapat, S., Barentsen, G., Bazán, J., Biswas, M., Boquien, M., Burke, D., Cara, D., Cara, M., Conroy, K., Conseil, S., Craig, M., Cross, R., Cruz, K., D'Eugenio, F., Dencheva, N., Devillepoix, H., Dietrich, J., Eigenbrot, A., Erben, T., Ferreira, L., Foreman-Mackey, D., Fox, R., Freij, N., Garg, S., Geda, R., Glattly, L., Gondhalekar, Y., Gordon, K., Grant, D., Greenfield, P., Groener, A., Guest, S., Gurovich, S., Handberg, R., Hart, A., Hatfield-Dodds, Z., Homeier, D., Hosseinzadeh, G., Jenness, T., Jones, C., Joseph, P., Kalmbach, J., Karamehmetoglu, E., Kałuszyński, M., Kelley, M., Kern, N., Kerzendorf, W., Koch, E., Kulumani, S., Lee, A., Ly, C., Ma, Z., MacBride, C., Maljaars, J., Muna, D., Murphy, N., Norman, H., O'Steen, R., Oman, K., Pacifici, C., Pascual, S., Pascual-Granado, J., Patil, R., Perren, G., Pickering, T., Rastogi, T., Roulston, B., Ryan, D., Rykoff, E., Sabater, J., Sakurikar, P., Salgado, J., Sanghi, A., Saunders, N., Savchenko, V., Schwardt, L., Seifert-Eckert, M., Shih, A., Jain, A., Shukla, G., Sick, J., Simpson, C., Singanamalla, S., Singer, L., Singhal, J., Sinha, M., Sipőcz, B., Spitler, L., Stansby, D., Streicher, O., Šumak, J., Swinbank, J., Taranu, D., Tewary, N., Tremblay, G., Val-Borro, M., Van Kooten, S., Vasović, Z., Verma, S., De Miranda Cardoso, J., Williams, P., Wilson, T., Winkel, B., Wood-Vasey, W., Xue, R., Yoachim, P., Zhang, C., Zonca, A.
+\& Astropy Project Contributors The Astropy Project: Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package.
+{\em The Astrophysical Journal}.
+\textbf{935}, e167 (2022,8)
 \end{thebibliography}

### 2025-02-01 paper/fig_5.py

@@ -12,7 +12,7 @@ def fig_5():
 	ys = (1/xs + 1/20) * times[0]
 	plt.plot(xs, ys, "r")
 
-	plt.scatter(ncpus[:1], times[:1], color="C3")
+	plt.scatter(ncpus[:1], times[:1], color="C3", marker="x")
 	plt.scatter(ncpus[1:], times[1:], color="C0")
 
 	plt.ylabel("Runtime [s]")

### 2025-02-01 paper/fig_7.py

@@ -1,0 +1,34 @@
+import matplotlib.pyplot as plt
+plt.style.use("timesnromanstyle.mplstyle")
+import numpy as np
+
+import sys
+sys.path.append("..")
+from scripts import arraytest
+
+
+def fig_7():
+	try:
+		mat = arraytest.run_experiment()
+	except FileNotFoundError:
+		print("Building figure 7 requires built parareal at parareal/build/parareal. Build it there using `mkdir parareal/build; cd parareal/build; cmake ../src; cmake --build .`", file=sys.stderr)
+	
+	fig, ax = plt.subplots()
+	ax.matshow(mat.T, cmap="Reds")
+	ax.set_xticks(np.arange(len(arraytest.COARSE_RANGE)))
+	ax.set_xticklabels(np.array(list(arraytest.COARSE_RANGE))/arraytest.FINE, rotation=45)
+	ax.set_yticks(np.arange(len(arraytest.EPSILONS)))
+	ax.set_yticklabels([f"$10^{{{-i}}}$" for i in arraytest.EPSILON_POWERS])
+	for (x, y), value in np.ndenumerate(mat):
+	    ax.text(x, y, str(value), va="center", ha="center")
+	ax.set_title("Required iterations for convergence")
+	ax.set_xlabel(r"$\frac{N_\mathcal{C}}{N_\mathcal{F}}$")
+	ax.set_ylabel(r"$\varepsilon$ (convergence threshold)")
+	plt.tick_params(axis="both", which="both", bottom=False, top=False, left=False, right=False, labeltop=False, labelbottom=True)
+	plt.minorticks_off()
+	plt.tight_layout()
+	plt.savefig("fig_7.eps", format="eps")
+
+
+if __name__ == "__main__":
+	fig_7()

### 2025-02-01 scripts/arraytest.py

@@ -16,41 +16,52 @@ TIME = 365
 FINE = 100
 
 COARSE_RANGE = range(10, 101, 10)
-EPSILON_POWERS = range(1, 6)
+EPSILON_POWERS = range(1, 10)
 EPSILONS = [10**(-n) for n in EPSILON_POWERS]
 
 
-mat = np.zeros((len(COARSE_RANGE), len(EPSILONS)), dtype=np.uint16)
-for i, coarse_steps in enumerate(COARSE_RANGE):
-	for j, eps in enumerate(EPSILONS):
-		result = subprocess.check_output([
-			PROGRAM_PATH, 
-			"-i", INPUT_PATH, 
-			"-o", "tmp.out", 
-			"--nseg", str(SEGMENTS),
-			"--time", str(TIME),
-			"--max-iters", str(SEGMENTS*2),
-			"--eps", str(eps),
-			"--coarse-steps", str(coarse_steps),
-			"--fine-steps", str(FINE)
-		])
-		iters = len(result.rstrip(b"\n").split(b"\n"))
-		print(f"{coarse_steps=} {eps=} {iters=}")
+def run_experiment():
+	mat = np.zeros((len(COARSE_RANGE), len(EPSILONS)), dtype=np.uint16)
+	for i, coarse_steps in enumerate(COARSE_RANGE):
+		for j, eps in enumerate(EPSILONS):
+			result = subprocess.check_output([
+				PROGRAM_PATH, 
+				"-i", INPUT_PATH, 
+				"-o", "tmp.out", 
+				"--nseg", str(SEGMENTS),
+				"--time", str(TIME),
+				"--max-iters", str(SEGMENTS*2),
+				"--eps", str(eps),
+				"--coarse-steps", str(coarse_steps),
+				"--fine-steps", str(FINE)
+			])
+			iters = len(result.rstrip(b"\n").split(b"\n"))
+			print(f"{coarse_steps=} {eps=} {iters=}")
 
-		mat[i, j] = iters
+			mat[i, j] = iters
 
-os.remove("tmp.out")
+	os.remove("tmp.out")
 
-fig, ax = plt.subplots()
-ax.matshow(mat.T, cmap="Reds")
-ax.set_xticks(np.arange(len(COARSE_RANGE)))
-ax.set_xticklabels(COARSE_RANGE)
-ax.set_yticks(np.arange(len(EPSILONS)))
-ax.set_yticklabels([f"$10^{{{-i}}}$" for i in EPSILON_POWERS])
[…]

### 2025-02-01 scripts/dist.py

@@ -4,11 +4,11 @@ import sys
 
 
 def compare(parareal_outfile, serial_outfile):
-	data_1 = np.loadtxt(sys.argv[1])
+	data_1 = np.loadtxt(parareal_outfile)
 	data_1, energy_1 = data_1[:, :-2], data_1[:, -2:]
 	data_1 = data_1.reshape((data_1.shape[0], data_1.shape[1]//3, 3))
 
-	data_2 = np.loadtxt(sys.argv[2])
+	data_2 = np.loadtxt(serial_outfile)
 	data_2, energy_2 = data_2[:, :-2], data_2[:, -2:]
 	data_2 = data_2.reshape((data_2.shape[0], data_2.shape[1]//3, 3))
 

### 2025-02-01 paper/fig_4.py
Fixed font in Figure 4

@@ -1,8 +1,8 @@
 import matplotlib.pyplot as plt
 plt.style.use("timesnromanstyle.mplstyle")
-plt.rcParams.update({
-	"text.usetex": True,
-})
+#plt.rcParams.update({
+#	"text.usetex": True,
+#})
 import numpy as np
 
 import sys
@@ -24,7 +24,7 @@ def fig_4():
 	ax.plot(np.arange(len(energy))*log_interval * dt / year, energy.sum(axis=1), label="total energy")
 	ax.legend()
 	plt.xlabel("Time [year]")
-	plt.ylabel(r"Energy $\left[\textrm{M}_{\textrm{Earth}}\frac{\textrm{AU}^2}{\textrm{day}^2}\right]$")
+	plt.ylabel(r"Energy $\left[\mathrm{M}_{\mathrm{Earth}}\frac{\mathrm{AU}^2}{\mathrm{day}^2}\right]$")
 
 	plt.tight_layout()
 	plt.savefig("fig_4.eps", format="eps")

### 2025-02-06 paper/main.tex
Fixed links
Paper improvements

@@
 \usepackage[thinc]{esdiff}
-\usepackage{xurl}
+\usepackage{hyperref}
 
@@
 \author{\IEEEauthorblockN{Lukáš Veškrna}
-\IEEEauthorblockA{\textit{Technical University Munich}\\ \textit{\& Charles University} \\
+\IEEEauthorblockA{\textit{Technical University Munich}\\ \textit{\& Charles University (Prague)} \\
 Munich, Germany \\
@@
 \begin{abstract}
-The purpose of this paper is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration.
+The purpose of this paper is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator for solar system dynamics that utilizes parallel-in-time integration.
 We present the results of experiments measuring the run times of the simulation, which we then compare with expectations derived from theory.
@@
 
-If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time. Thus, parallel-in-time algorithms can be employed. \cite{parallelsolar}
+To study the behavior of mostly stable planetary orbits, long simulations are required. Thus, parallel-in-time algorithms provide a promising approach. \cite{parallelsolar}
 
@@
 We typically solve this problem using a numerical method, that does a step in time of size $\Delta t$ and estimates $u(t_0+\Delta t)$.
-We then use this result to advance the computation further in time by taking more such steps.
+We then use this result to advance the computation further in time by taking more such steps.\footnote{One such method is the explicit Euler method, which simply approximates $u$ at time $t+\Delta t$ by taking a step in the direction given by the value of $f$ at the current state $u$ at time $t$, so $u(t+\Delta t) \approx u(t) + \Delta t f(t, u(t))$.
+\cite{forrk4}}
 
@@
 
-Supposing we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta t = \frac{T-t_0}{N}$.
+Suppose we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta t = \frac{T-t_0}{N}$.
 We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta t$ (so $t_0 = t_0$ and $t_N = T$).\footnote{The solvers can also use their own smaller subdivisions (e.g.
@@
 This is repeated until satisfactory convergence.
-It has been proven for example by \cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta t, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
+It has been proven for example by Gander et al.
+\cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta t, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.
 
-This method has been formulated using different generalized frameworks in \cite{parareal2}.
-They also extend the time interval paritioning to arbitrary partions, not only equally sized ones as in \cite{parareal}.
+In the work by Gander et al.
+\cite{parareal2}, the method has been formulated using different generalized frameworks.
+They also extend the time interval partitioning to arbitrary partitions, not only equally sized ones as in the original paper.
+\cite{parareal}
 
@@
 \cite{speedup} (Section 2.3), which we use here.
-It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially (our implementation doesn't include this optimization).
+It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it by running $\mathcal{C}$ serially (our implementation doesn't include this optimization).
 
@@
 This was done prior to the invention of parareal by Saha et al.
-\cite{parallelsolar}, also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.
+\cite{parallelsolar}, also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of the Sun and other planets.
 \cite{parallelsolar} \cite{farfuture}
@@
 \cite{integration}
-    \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic.
+    \item Two symplectic methods could be used\footnote{A symplectic integrator preserves physically meaningful behavior of a Hamiltonian system in the long term.
+\cite{integration}}, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic.
 A modified parareal algorithm that preserves the symplectic property, shown in \cite{symplecticparareal}, also exists, however it is beyond the scope of this paper.
     \item An adaptive step size fine integrator with error control would be an option, such as the Dormand-Prince RK45 method.
-However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
+However, since the method can have very different execution times (depending on the step size required to reach the desired accuracy), which could potentially be an issue when parallelizing the code, RK4 was used for simplicity.
 \end{itemize}
@@
 \begin{align*}
-x(t+\Delta t) &= x(t) + v(t)\Delta t + \frac{1}{2}a(t)\Delta t^2 \\
-v(t+\Delta t) &= v(t) + \frac{a(t) + a(t+\Delta t)}{2}\Delta t
+x(t+\Delta t) &\approx x(t) + v(t)\Delta t + \frac{1}{2}a(t)\Delta t^2 \\
+v(t+\Delta t) &\approx v(t) + \frac{a(t) + a(t+\Delta t)}{2}\Delta t
 \end{align*}
-Here $x(t)$, $v(t)$ and $a(t)$ represent approximations of the corresponding values (position, velocity, acceleration) at timestep $t$, $\Delta t$ is the time step size.
+Here $x(t)$, $v(t)$ and $a(t)$ are approximations of the corresponding values (position, velocity, acceleration) by the method at time $t$, $\Delta t$ is the time step size.
 Note that only one evaluation of $a$ per step suffices, because we can keep $a(t)$ from the last step.
@@
 k_4 &= f(t + \Delta t, u(t) + \Delta t k_3) \\
-u(t+\Delta t) &= u(t) + \frac{\Delta t}{6}\left(k_1 + 2k_2 + 2k_3 + k_4\right)
+u(t+\Delta t) &\approx u(t) + \frac{\Delta t}{6}\left(k_1 + 2k_2 + 2k_3 + k_4\right)
 \end{align*}
-where $u(t)$ represents the approximation of the state at time $t$. $\Delta t$ denotes the time step size. The method requires four acceleration computations per step. \cite{forrk4}
+where $u(t)$ is the approximation of the state by the method at time $t$. $\Delta t$ denotes the time step size. The method requires four acceleration computations per step. \cite{forrk4}
 
@@
 \caption{A more complex example to illustrate the behavior of the parareal algorithm: Path of Mercury over the time of one Earth year in a parareal integrated simulation converging over multiple iterations of the parareal algorithm.
-Velocity verlet and RK4 are used with the same step size of $0.5\text{ day}$.}
+Velocity Verlet and RK4 are used with the same step size of $0.5\text{ day}$.}
 \label{mercury}
@@
 \subsection{Experiments: The environment}
-All experiments were ran on a cluster in order to get a more reproducible and comparable environment.
+All experiments were run on a cluster in order to get a more reproducible and comparable environment.
 
@@
 
-Multiple experiments with the same initial conditions and same step size were conducted.
+Multiple experiments with the same initial conditions and the same step size were conducted.
 A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration\footnote{This choice is explained in the next experiment (\textit{Experiment II}), where the observations suggest that convergence within one iteration possibly has optimal run time.}).
-For comparison, also a serial version (not using the parareal algorithm) was ran with the same initial conditions and step size.
+For comparison, also a serial version (not using the parareal algorithm) was run with the same initial conditions and step size.
 
@@
 Same simulation as in Figure \ref{sim}.
-With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of timesteps.}
+With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of time steps.}
 \label{energy}
@@
 
-Let's derive the expected speed-ups and run times of our program and compare it with the measured run times.
+Let's derive the expected speed-ups and run times of our program and compare them with the measured run times.
 To simplify the calculation, we will only estimate it using the number of function evaluations.
@@
 We can compare this with the real measured speed-up, see Table \ref{runtimes}.
-We observe that in reality the program runs around 5 times faster with 16 threads.
+We observe that in reality, the program runs around 5 times faster with 16 threads.
 We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure \ref{speeds}.
@@
 
-While we loose some accuracy with the parareal method, this is a parameter that can be controlled, for example using the converge parameter ($\varepsilon$) and the coarse solver accuracy.
+While we lose some accuracy with the parareal method, this is a parameter that can be controlled, for example using the converge parameter ($\varepsilon$) and the coarse solver accuracy.
 
@@
 
-The second experiment shows how do run times and the numbers of iterations that the program takes to converge relate to the number of steps of the coarse solver.
+The second experiment shows how run times and the numbers of iterations that the program takes to converge relate to the number of steps of the coarse solver.
 This observation can help with deciding what is the optimal number of steps to use for the coarse solver.
@@
 
-A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were ran and the results are shown in Figure \ref{timetest}.
+A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were run and the results are shown in Figure \ref{timetest}.
 The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.
@@
 
-From this, we can conjecture, that (in the case of our simulation) the optimal choice of the number of steps for the coarse solver (in the sense of the smallest run time with a fixed $\varepsilon$) is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).
+From this, we can conjecture, that (in the case of our simulation) the optimal choice of the number of steps for the coarse solver (in the sense of the smallest run time with a fixed $\varepsilon$) is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, the optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).
 
@@
 $N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively (within one segment).
-The fine solver step numbers have been chosen specifically to preserve the step size of 0.003125 day.
+The fine solver step numbers have been chosen specifically to preserve the step size of 0.003125 days.
 This step size was selected, so all the parareal runs converge within one iteration (with $\varepsilon = 10^{-3}\text{ AU}$).
 This number of iterations was chosen for easier comparison between the runs and to achieve greater speed-up (see \textit{Experiment II}).
-The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
+The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equipped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
 \begin{center}
@@
 
-The implementation is consistently slower than the estimated run times, however, there are many factors that weren't considered in the speed-up estimation, for example we only took the function evaluations in the solvers into account.
+The implementation is consistently slower than the estimated run times; however, many factors were not considered in the speed-up estimation, for example, in the solvers we only took into account the function evaluations.
 There are also other factors such as the synchronization costs of the OpenMP threads.
 
-In conclusion, we were able to achieve noticable speed-ups consistent with the theoretical foundation by using the parareal algorithm.
+In conclusion, we were able to achieve noticeable speed-ups consistent with the theoretical foundation by using the parareal algorithm.
 
@@
 
-And finally in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation.
+Finally, in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation.
 The results are presented in Figure \ref{timetest}.
@@
 
-In some situations it is not that clear as in our case whether is it better to use spatial or temporal parallelization or both.
+In some situations, it is not that clear as in our case whether is it better to use spatial or temporal parallelization or both.
 This is beyond the scope of this article, however, there has been research in this area.
@@
 presented a study using a simulation of unsteady Navier-Stokes equations for incompressible flow using parareal that is parallel both in time and space.
-They measured the speed-up with different numbers of cores being asigned to the spatial and temporal parallelization and shown that they can be combined to gain higher speed-up.
+They measured the speed-up with different numbers of cores assigned to the spatial and temporal parallelization and showed that they can be combined to gain higher speed-up.
 \cite{spacetime}
@@
 We evaluated parareal only in the example of our solar system.
-Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate in the sense that if we simulate a single orbit of Neptune, we have to simulate many orbits of Mercury (in a simulation of all the planets), which requires very small step sizes to behave correctly.
+Those settings, however, have their own specifics, for example, planets in our solar system follow orbits with very different periods, which makes them harder to simulate in the sense that if we simulate a single orbit of Neptune, we have to simulate many orbits of Mercury (in a simulation of all the planets), which requires very small step sizes to behave correctly.
 
 We could further improve our solar system simulations by further specializing them as was done in the work by Saha et al.
-\cite{parallelsolar} (a different coarse model simulating only the interactions of Sun and other planets was used instead of a course integrator).
+\cite{parallelsolar} (a different coarse model simulating only the interactions of the Sun and other planets was used instead of a course integrator).
 However, this would be too specific for solar system dynamics and wound not generalize well, so this is beyond our scope.
@@
 Sc.
-and all other participants of the Seminar in Methods for Molecular Dynamics, for suggesting directions, providing valueable feedback and organising this amazing seminar.
+and all other participants of the Seminar in Methods for Molecular Dynamics, for suggesting directions, providing valuable feedback and organizing this amazing seminar.
 
@@
 
-\IEEEtriggeratref{9}
+\IEEEtriggeratref{7}
 \begin{thebibliography}{00}
@@
 \textbf{332}, 661-668 (2001), [Online].
-Available: \url{https://www.sciencedirect.com/science/article/pii/S0764444200017936}
+Available: \url{https://doi.org/10.1016/S0764-4442(00)01793-6}
 \bibitem{parallelsolar}Saha, P., Stadel, J.
@@
 21 (2019), [Online].
-Available: \url{http://publications.mfo.de/handle/mfo/3693}
+Available: \url{https://doi.org/10.14760/SNAP-2019-021-EN}
 \bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)

### 2025-02-07 paper/main.tex
Improved abstract

@@
 \begin{abstract}
-The purpose of this paper is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator for solar system dynamics that utilizes parallel-in-time integration.
-We present the results of experiments measuring the run times of the simulation, which we then compare with expectations derived from theory.
+This paper examines parallel-in-time integration, specifically the parareal algorithm, for particle simulations by developing a simple parallel-in-time simulator for solar system dynamics.
+We present experimental measurements of the simulation run times, which we then compare with expectations derived from theory.
+Additionally, we explore the behavior of the algorithm with different parameter settings.
 \end{abstract}

### 2025-02-07 paper/main.tex
Line breaks at figure refs

@@
 
-The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figure \ref{sim} and \ref{energy}.
+The resulting run times were collected in Table~\ref{runtimes}. For visualizations of the simulation results see Figures~\ref{sim}~and~\ref{energy}.
 
@@
 \centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
-\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal. For more information, see Table \ref{runtimes}.}
+\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal. For more information, see Table~\ref{runtimes}.}
 \label{sim}
@@
 \caption{Conservation of energy in the simulation.
-Same simulation as in Figure \ref{sim}.
+Same simulation as in Figure~\ref{sim}.
 With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of time steps.}
@@
 
-We can compare this with the real measured speed-up, see Table \ref{runtimes}.
+We can compare this with the real measured speed-up, see Table~\ref{runtimes}.
 We observe that in reality, the program runs around 5 times faster with 16 threads.
-We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure \ref{speeds}.
+We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure~\ref{speeds}.
 
@@
 \caption{Comparison of the expected run times (red curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red ``x'' point)) and the real measured times with different numbers of CPU cores (blue points).
-For more information, see Table \ref{runtimes}.}
+For more information, see Table~\ref{runtimes}.}
 \label{speeds}
@@
 If we instead approximate the simulation using a similar serial simulation with a lower number of steps, we get a substantially bigger error than the parareal algorithm while not improving the run time significantly.
-This is also shown in Table \ref{runtimes} (as ``serial (approx.)'').
+This is also shown in Table~\ref{runtimes} (as ``serial (approx.)'').
 
@@
 
-A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were run and the results are shown in Figure \ref{timetest}.
+A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were run and the results are shown in Figure~\ref{timetest}.
 The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.
 
-A key feature observed in Figure \ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes.
+A key feature observed in Figure~\ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes.
 After every initial downward spike, there is always an upward trend.
@@
 The run times of a fixed simulation with different numbers of CPU cores were measured in Experiment I and compared with the theoretical expectations.
-As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).
+As was visualized in Figure~\ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).
 
@@
 
-By comparing parareal with an approximating serial run using a bigger step size (see ''serial (approx.)`` in Table \ref{runtimes}), we have also shown that in this case, the parareal algorithm beats simply approximating the solution using a smaller step size in both accuracy and run time.
+By comparing parareal with an approximating serial run using a bigger step size (see ''serial (approx.)`` in Table~\ref{runtimes}), we have also shown that in this case, the parareal algorithm beats simply approximating the solution using a smaller step size in both accuracy and run time.
 
@@
 Finally, in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation.
-The results are presented in Figure \ref{timetest}.
+The results are presented in Figure~\ref{timetest}.
 The result is a prediction of the optimal coarse solver step size for our simulation, which is the biggest possible such that the algorithm converges in one iteration.
